---
title: Suyash's project
engine: knitr
format:
  html:
    df-print: kable
    include-in-header:
      - text: |
          <script src="../js/dagitty.js"></script>
---

```{r project_config, echo=FALSE, warning=FALSE, message=FALSE}
variables <- c(
  WST = 'ER.H2O.FWST.ZS',
  DWS = 'SH.H2O.BASW.ZS',
  RUR = 'SP.RUR.TOTL.ZS',
  WSW = 'SL.EMP.WORK.ZS' # WSW
)

histograms <- list(
  c(variables['DWS'], variables['WST'])
)

ALPHA_MEAN <- 50
ALPHA_SD <- 25
BETA_SD <- 10
SIGMA_SD <- 20

dag <- '
dag {
bb="0,0,1,1"
"People with Drinking Water Services" [exposure,pos="0.168,0.545"]
"Rural Population (%)" [adjusted,pos="0.520,0.364"]
"Wage/Salaried Workers" [adjusted,pos="0.519,0.207"]
"Water Stress" [outcome,pos="0.840,0.550"]
"People with Drinking Water Services" -> "Water Stress"
"Rural Population (%)" -> "People with Drinking Water Services"
"Rural Population (%)" -> "Water Stress"
"Wage/Salaried Workers" -> "People with Drinking Water Services"
"Wage/Salaried Workers" -> "Water Stress"
}

'
```

{{< include _base.qmd >}}


# Formal Model
$$
\begin{align}
WST_i &\sim \text{Normal}(\mu_i,\sigma) \\
\mu_i &= \alpha
+	\beta_{DWS} DWS_i
+	\beta_{RUR} RUR_i
+	\beta_{WSW} WSW_i \\
\alpha &\sim \text{Normal}(`r ALPHA_MEAN`,`r ALPHA_SD`) \\
\beta_{DWS} &\sim \text{Normal}(0,`r BETA_SD`) \\
\beta_{RUR} &\sim \text{Normal}(0,`r BETA_SD`) \\
\beta_{WSW} &\sim \text{Normal}(0,`r BETA_SD`) \\
\sigma &\sim \text{Normal}(0,`r SIGMA_SD`), \quad \sigma > 0
\end{align}
$$

| Variable | Description | Unit |
|----------|-------------|------|
| $WST_i$  | Water Stress (draw vs availability) | percentage |
| $DWS_i$  | Population with Drinking Water Service (%) | standardized |
| $RUR_i$  | Rural Population (percent) | standardized |
| $WSW_i$  | Government Effectiveness (percentile) | standardized |


# Base model

Just $WST_i$ ~ $DWS_i$

```{stan model1_stan, eval=FALSE, output.var="model1_stan"}
data {
  int<lower=1> n;
  vector[n] WST;
  vector[n] DWS;
}

parameters {
  real alpha;
  real beta_DWS;
  real<lower=0> sigma;
}

transformed parameters {
  vector[n] mu = alpha + beta_DWS * DWS;
}

model {
  alpha    ~ normal(<<ALPHA_MEAN>>, <<ALPHA_SD>>);
  beta_DWS ~ normal(0, <<BETA_SD>>);
  sigma    ~ normal(0, <<SIGMA_SD>>);

  WST ~ normal(mu, sigma);
}

generated quantities {
  vector[n] WST_rep;
  for (i in 1:n) {
    WST_rep[i] = normal_rng(mu[i], sigma);
  }
}
```

```{r model1, message=FALSE, results="hide"}
model1 <- knitr::knit_code$get("model1_stan") |>
  paste(collapse = "\n") |>
  glue::glue(.open = "<<", .close = ">>") |>
  write_stan_file() |>
  cmdstan_model()

fit1 <- df |>
  transmute(
    WST = .data[[variables['WST']]],
    DWS = .data[[variables['DWS']]] |> scale() |> as.numeric(),
  ) |>
  drop_na() |>
  compose_data() |>
  model1$sample()
```

```{r}
fit1$summary() |>
  filter(
    !stringr::str_detect(variable, "mu\\[") &
      !stringr::str_detect(variable, "_rep\\[") &
      !stringr::str_detect(variable, "lp__")
  )

draws1 <- fit1 |>
  spread_draws(alpha, beta_DWS, sigma)
plot_param(draws1, "beta_DWS")
```

# Statistical model

Impact of Water Services on Water Stress

Red is our prior distribution; blue is our posterior distribution.

```{stan model2_stan, eval=FALSE, output.var="model2_stan"}
data {
  int<lower=1> n;
  vector[n] WST;
  vector[n] DWS;
  vector[n] RUR;
  vector[n] WSW;
}

parameters {
  real alpha;
  real beta_DWS;
  real beta_RUR;
  real beta_WSW;
  real<lower=0> sigma;
}

transformed parameters {
  vector[n] mu = alpha
               + beta_DWS * DWS
               + beta_RUR * RUR
               + beta_WSW * WSW;
}

model {
  alpha    ~ normal(<<ALPHA_MEAN>>, <<ALPHA_SD>>);
  beta_DWS  ~ normal(0, <<BETA_SD>>);
  beta_RUR  ~ normal(0, <<BETA_SD>>);
  beta_WSW  ~ normal(0, <<BETA_SD>>);
  sigma    ~ normal(0, <<SIGMA_SD>>);

  WST ~ normal(mu, sigma);
}

generated quantities {
  vector[n] WST_rep;
  for (i in 1:n) {
    WST_rep[i] = normal_rng(mu[i], sigma);
  }
}
```

```{r model2, message=FALSE, results="hide"}
model2 <- knitr::knit_code$get("model2_stan") |>
  paste(collapse = "\n") |>
  glue::glue(.open = "<<", .close = ">>") |>
  write_stan_file() |>
  cmdstan_model()

fit2 <- df |>
  transmute(
    WST = .data[[variables['WST']]],
    DWS = .data[[variables['DWS']]] |> scale() |> as.numeric(),
    RUR = .data[[variables['RUR']]] |> scale() |> as.numeric(),
    WSW = .data[[variables['WSW']]] |> scale() |> as.numeric()
  ) |>
  drop_na() |>
  compose_data() |>
  model2$sample()
```

```{r}
fit2$summary() |>
  filter(
    !stringr::str_detect(variable, "mu\\[") &
      !stringr::str_detect(variable, "_rep\\[") &
      !stringr::str_detect(variable, "lp__")
  )

params <- c("beta_DWS", "beta_RUR", "beta_WSW")
draws <- fit2 |>
  spread_draws(alpha, !!!syms(params), sigma)
params |> purrr::walk(\(param) print(plot_param(draws, param)))
```