---
title: Suyash's project
engine: knitr
format:
  html:
    df-print: kable
    include-in-header:
      - text: |
          <script src="../js/dagitty.js"></script>
---

```{r project_config, echo=FALSE, warning=FALSE, message=FALSE}
variables <- c(
  PSB = 'PV.PER.RNK',
  COR = 'CC.PER.RNK',
  VNA = 'VA.PER.RNK',
  GEF = 'GE.PER.RNK',
  ROL = 'RL.PER.RNK' #ROL
)

histograms <- list(
  c(variables['COR'], variables['PSB'])
)

ALPHA_MEAN <- 50
ALPHA_SD <- 25
BETA_SD <- 10
SIGMA_SD <- 20

dag <- '
dag {
bb="0,0,1,1"
"Government Effectiveness" [adjusted,pos="0.510,0.318"]
"Political Stability" [outcome,pos="0.847,0.643"]
"Rule of Law" [adjusted,pos="0.514,0.193"]
"Voice and Accountability" [adjusted,pos="0.506,0.466"]
Corruption [exposure,pos="0.099,0.642"]
"Government Effectiveness" -> "Political Stability"
"Government Effectiveness" -> Corruption
"Rule of Law" -> "Political Stability"
"Rule of Law" -> Corruption
"Voice and Accountability" -> "Political Stability"
"Voice and Accountability" -> Corruption
Corruption -> "Political Stability"
}
'
```

{{< include _base.qmd >}}


# Formal Model
$$
\begin{align}
PSB_i &\sim \text{Normal}(\mu_i,\sigma) \\
\mu_i &= \alpha
+	\beta_{COR} COR_i
+	\beta_{VNA} VNA_i
+	\beta_{GEF} GEF_i
+	\beta_{ROL} ROL_i \\
\alpha &\sim \text{Normal}(`r ALPHA_MEAN`,`r ALPHA_SD`) \\
\beta_{COR} &\sim \text{Normal}(0,`r BETA_SD`) \\
\beta_{VNA} &\sim \text{Normal}(0,`r BETA_SD`) \\
\beta_{GEF} &\sim \text{Normal}(0,`r BETA_SD`) \\
\beta_{ROL} &\sim \text{Normal}(0,`r BETA_SD`) \\
\sigma &\sim \text{Normal}(0,`r SIGMA_SD`), \quad \sigma > 0
\end{align}
$$

| Variable | Description | Unit |
|----------|-------------|------|
| $PSB_i$  | Political stability | percentile |
| $COR_i$  | Perception of Corruption | standardized |
| $VNA_i$  | Voice and Accountability | standardized |
| $GEF_i$  | Government Effectiveness (percentile) | standardized |
| $ROL_i$  | Rule of Law | standardized |


# Base model

Just $PSB_i$ ~ $COR_i$

```{stan model1_stan, eval=FALSE, output.var="model1_stan"}
data {
  int<lower=1> n;
  vector[n] PSB;
  vector[n] COR;
}

parameters {
  real alpha;
  real beta_COR;
  real<lower=0> sigma;
}

transformed parameters {
  vector[n] mu = alpha + beta_COR * COR;
}

model {
  alpha    ~ normal(<<ALPHA_MEAN>>, <<ALPHA_SD>>);
  beta_COR ~ normal(0, <<BETA_SD>>);
  sigma    ~ normal(0, <<SIGMA_SD>>);

  PSB ~ normal(mu, sigma);
}

generated quantities {
  vector[n] PSB_rep;
  for (i in 1:n) {
    PSB_rep[i] = normal_rng(mu[i], sigma);
  }
}
```

```{r model1, message=FALSE, results="hide"}
model1 <- knitr::knit_code$get("model1_stan") |>
  paste(collapse = "\n") |>
  glue::glue(.open = "<<", .close = ">>") |>
  write_stan_file() |>
  cmdstan_model()

fit1 <- df |>
  transmute(
    PSB = .data[[variables['PSB']]],
    COR = .data[[variables['COR']]] |> scale() |> as.numeric(),
  ) |>
  drop_na() |>
  compose_data() |>
  model1$sample()
```

```{r}
fit1$summary() |>
  filter(
    !stringr::str_detect(variable, "mu\\[") &
      !stringr::str_detect(variable, "_rep\\[") &
      !stringr::str_detect(variable, "lp__")
  )

draws1 <- fit1 |>
  spread_draws(alpha, beta_COR, sigma)
plot_param(draws1, "beta_COR")
```

# Statistical model

Impact of Corruption on Political Stability

Red is our prior distribution; blue is our posterior distribution.

```{stan model2_stan, eval=FALSE, output.var="model2_stan"}
data {
  int<lower=1> n;
  vector[n] PSB;
  vector[n] COR;
  vector[n] VNA;
  vector[n] GEF;
  vector[n] ROL;
}

parameters {
  real alpha;
  real beta_COR;
  real beta_VNA;
  real beta_GEF;
  real beta_ROL;
  real<lower=0> sigma;
}

transformed parameters {
  vector[n] mu = alpha
               + beta_COR * COR
               + beta_VNA * VNA
               + beta_GEF * GEF
               + beta_ROL * ROL;
}

model {
  alpha    ~ normal(<<ALPHA_MEAN>>, <<ALPHA_SD>>);
  beta_COR  ~ normal(0, <<BETA_SD>>);
  beta_VNA  ~ normal(0, <<BETA_SD>>);
  beta_GEF  ~ normal(0, <<BETA_SD>>);
  beta_ROL  ~ normal(0, <<BETA_SD>>);
  sigma    ~ normal(0, <<SIGMA_SD>>);

  PSB ~ normal(mu, sigma);
}

generated quantities {
  vector[n] PSB_rep;
  for (i in 1:n) {
    PSB_rep[i] = normal_rng(mu[i], sigma);
  }
}
```

```{r model2, message=FALSE, results="hide"}
model2 <- knitr::knit_code$get("model2_stan") |>
  paste(collapse = "\n") |>
  glue::glue(.open = "<<", .close = ">>") |>
  write_stan_file() |>
  cmdstan_model()

fit2 <- df |>
  transmute(
    PSB = .data[[variables['PSB']]],
    COR = .data[[variables['COR']]] |> scale() |> as.numeric(),
    VNA = .data[[variables['VNA']]] |> scale() |> as.numeric(),
    GEF = .data[[variables['GEF']]] |> scale() |> as.numeric(),
    ROL = .data[[variables['ROL']]] |> scale() |> as.numeric()
  ) |>
  drop_na() |>
  compose_data() |>
  model2$sample()
```

```{r}
fit2$summary() |>
  filter(
    !stringr::str_detect(variable, "mu\\[") &
      !stringr::str_detect(variable, "_rep\\[") &
      !stringr::str_detect(variable, "lp__")
  )

params <- c("beta_COR", "beta_VNA", "beta_GEF", "beta_ROL")
draws <- fit2 |>
  spread_draws(alpha, !!!syms(params), sigma)
params |> purrr::walk(\(param) print(plot_param(draws, param)))
```