---
title: Sambit's project
engine: knitr
format:
  html:
    df-print: kable
    include-in-header:
      - text: |
          <script src="../js/dagitty.js"></script>
---

```{r project_config, echo=FALSE, warning=FALSE, message=FALSE}
variables <- c(
  YN = 'SL.UEM.NEET.ME.ZS',
  IN = 'IT.NET.USER.ZS',
  EU = 'SP.URB.TOTL.IN.ZS',
  EA = 'EG.ELC.ACCS.ZS',
  UR = 'NY.GDP.PCAP.PP.CD'
)

histograms <- list(
  c('IT.NET.USER.ZS', 'SL.UEM.NEET.ME.ZS')
)

ALPHA_MEAN <- 50
ALPHA_SD <- 25
BETA_SD <- 5
SIGMA_SD <- 20

dag <- '
dag {
bb="0,0,1,1"
"Electricity Access" [adjusted,pos="0.560,0.288"]
"GDP per Capita" [adjusted,pos="0.195,0.309"]
"Internet Users (%)" [exposure,pos="0.190,0.777"]
"Youth NEETs" [outcome,pos="0.878,0.802"]
Urbanization [adjusted,pos="0.867,0.357"]
"Electricity Access" -> "Internet Users (%)"
"Electricity Access" -> "Youth NEETs"
"GDP per Capita" -> "Electricity Access"
"GDP per Capita" -> "Internet Users (%)"
"GDP per Capita" -> "Youth NEETs"
"Internet Users (%)" -> "Youth NEETs"
Urbanization -> "Electricity Access"
Urbanization -> "Youth NEETs"
}
'
```

{{< include _base.qmd >}}


# Formal Model
$$
\begin{align}
YN_i &\sim \text{Normal}(\mu_i,\sigma) \\
\mu_i &= \alpha
+	\beta_{IU}  IU_i
+	\beta_{IN} IN_i
+	\beta_{EA} EA_i
+	\beta_{UR} UR_i \\
\alpha &\sim \text{Normal}(`r ALPHA_MEAN`,`r ALPHA_SD`) \\
\beta_{IU} &\sim \text{Normal}(0,`r BETA_SD`) \\
\beta_{IN} &\sim \text{Normal}(0,`r BETA_SD`) \\
\beta_{EA} &\sim \text{Normal}(0,`r BETA_SD`) \\
\beta_{UR} &\sim \text{Normal}(0,`r BETA_SD`) \\
\sigma &\sim \text{Normal}(0,`r SIGMA_SD`), \quad \sigma > 0
\end{align}
$$

| Variable | Description | Unit |
|----------|-------------|------|
| $YN_i$  | Youth not in Education, Employment, or Training | percentage |
| $IU_i$  | Internet Users (%) | standardized |
| $IN_i$  | GDP, per capita (PPP) | standardized |
| $EA_i$  | Electricity Access (%) | standardized |
| $UR_i$  | Urbanization (%) | standardized |


# Base model

Just $YN_i$ ~ $IU_i$

```{stan model1_stan, eval=FALSE, output.var="model1_stan"}
data {
  int<lower=1> n;
  vector[n] YN;
  vector[n] IU;
}

parameters {
  real alpha;
  real beta_IU;
  real<lower=0> sigma;
}

transformed parameters {
  vector[n] mu = alpha + beta_IU * IU;
}

model {
  alpha    ~ normal(<<ALPHA_MEAN>>, <<ALPHA_SD>>);
  beta_IU ~ normal(0, <<BETA_SD>>);
  sigma    ~ normal(0, <<SIGMA_SD>>);

  YN ~ normal(mu, sigma);
}

generated quantities {
  vector[n] YN_rep;
  for (i in 1:n) {
    YN_rep[i] = normal_rng(mu[i], sigma);
  }
}
```

```{r model1, message=FALSE, results="hide"}
model1 <- knitr::knit_code$get("model1_stan") |>
  paste(collapse = "\n") |>
  glue::glue(.open = "<<", .close = ">>") |>
  write_stan_file() |>
  cmdstan_model()

fit1 <- df |>
  transmute(
    YN = SL.UEM.NEET.ME.ZS,
    IU = IT.NET.USER.ZS |> scale() |> as.numeric(),
  ) |>
  drop_na() |>
  compose_data() |>
  model1$sample()
```

```{r}
fit1$summary() |>
  filter(
    !stringr::str_detect(variable, "mu\\[") &
      !stringr::str_detect(variable, "_rep\\[") &
      !stringr::str_detect(variable, "lp__")
  )

draws1 <- fit1 |>
  spread_draws(alpha, beta_IU, sigma)
plot_param(draws1, "beta_IU")
```

# Statistical model

Impact of Internet Access on Youth NEETs

Red is our prior distribution; blue is our posterior distribution.

```{stan model2_stan, eval=FALSE, output.var="model2_stan"}
data {
  int<lower=1> n;
  vector[n] YN;
  vector[n] IU;
  vector[n] IN;
  vector[n] EA;
  vector[n] UR;
}

parameters {
  real alpha;
  real beta_IU;
  real beta_IN;
  real beta_EA;
  real beta_UR;
  real<lower=0> sigma;
}

transformed parameters {
  vector[n] mu = alpha
               + beta_IU * IU
               + beta_IN * IN
               + beta_EA * EA
               + beta_UR * UR;
}

model {
  alpha    ~ normal(<<ALPHA_MEAN>>, <<ALPHA_SD>>);
  beta_IU  ~ normal(0, <<BETA_SD>>);
  beta_IN  ~ normal(0, <<BETA_SD>>);
  beta_EA  ~ normal(0, <<BETA_SD>>);
  beta_UR  ~ normal(0, <<BETA_SD>>);
  sigma    ~ normal(0, <<SIGMA_SD>>);

  YN ~ normal(mu, sigma);
}

generated quantities {
  vector[n] YN_rep;
  for (i in 1:n) {
    YN_rep[i] = normal_rng(mu[i], sigma);
  }
}
```

```{r model2, message=FALSE, results="hide"}
model2 <- knitr::knit_code$get("model2_stan") |>
  paste(collapse = "\n") |>
  glue::glue(.open = "<<", .close = ">>") |>
  write_stan_file() |>
  cmdstan_model()

fit2 <- df |>
  transmute(
    YN = SL.UEM.NEET.ME.ZS,
    IU = IT.NET.USER.ZS |> scale() |> as.numeric(),
    IN = SP.URB.TOTL.IN.ZS |> scale() |> as.numeric(),
    EA = EG.ELC.ACCS.ZS |> scale() |> as.numeric(),
    UR = NY.GDP.PCAP.PP.CD |> scale() |> as.numeric()
  ) |>
  drop_na() |>
  compose_data() |>
  model2$sample()
```

```{r}
fit2$summary() |>
  filter(
    !stringr::str_detect(variable, "mu\\[") &
      !stringr::str_detect(variable, "_rep\\[") &
      !stringr::str_detect(variable, "lp__")
  )

params <- c("beta_IU", "beta_IN", "beta_EA", "beta_UR")
draws <- fit2 |>
  spread_draws(alpha, !!!syms(params), sigma)
params |> purrr::walk(\(param) print(plot_param(draws, param)))
```