---
title: Sameer's project
engine: knitr
format:
  html:
    df-print: kable
    include-in-header:
      - text: |
          <script src="../js/dagitty.js"></script>
---

```{r project_config, echo=FALSE, warning=FALSE, message=FALSE}
variables <- c(
  'SL.EMP.VULN.ZS',
  'SE.COM.DURS',
  'SL.UEM.NEET.ME.ZS',
  'RL.EST',
  'VA.EST'
)

histograms <- list(
  c('SE.COM.DURS', 'SL.EMP.VULN.ZS')
)

dag <- '
"Compulsory Education Duration"[exposure, pos = "-0.917,0.137"]
"Rule of Law"[adjusted, pos = "-0.385,-0.178"]
"Voice and Accountability"[adjusted, pos = "-0.375,-0.477"]
"Vulnerable Employment"[outcome, pos = "0.169,0.153"]
"Youth NEETs"[pos = "-0.375,0.448"]
"Compulsory Education Duration" -> "Vulnerable Employment"
"Compulsory Education Duration" -> "Youth NEETs"
"Rule of Law" -> "Compulsory Education Duration"
"Rule of Law" -> "Vulnerable Employment"
"Voice and Accountability" -> "Compulsory Education Duration"
"Voice and Accountability" -> "Vulnerable Employment"
"Youth NEETs" -> "Vulnerable Employment"
'
```

{{< include _base.qmd >}}


# Formal Model
$$
\begin{align}
VE_i &\sim \text{Normal}(\mu_i,\sigma) \\
\mu_i &= \alpha
+	\beta_{CED}  CED_i
+	\beta_{YN} YN_i
+	\beta_{RL} RL_i
+	\beta_{VA} VA_i \\
\alpha &\sim \text{Normal}(50,20) \\
\beta_{CED} &\sim \text{Normal}(0,1) \\
\beta_{YN} &\sim \text{Normal}(0,1) \\
\beta_{RL} &\sim \text{Normal}(0,1) \\
\beta_{VA} &\sim \text{Normal}(0,1) \\
\sigma &\sim \text{Normal}(0,10), \quad \sigma > 0
\end{align}
$$

| Variable | Description | Unit |
|----------|-------------|------|
| $VE_i$  | Vulnerable Employment | percentage |
| $CED_i$  | Compulsory Education Duration | standardized |
| $YN_i$  | Youth not in Education, Employment, or Training | standardized |
| $RL_i$  | Rule of Law | standardized |
| $VA_i$  | Voice and Accountability | standardized |


# Statistical model

Impact of different variables on Vulnerable Employment rates

Red is our prior distribution; blue is our posterior distribution.
```{r model, message=FALSE, results="hide"}
model <- write_stan_file(
  "
data {
  int<lower=1> n;
  vector[n] VE;
  vector[n] CED;
  vector[n] YN;
  vector[n] RL;
  vector[n] VA;
}

parameters {
  real alpha;
  real beta_CED;
  real beta_YN;
  real beta_RL;
  real beta_VA;
  real<lower=0> sigma;
}

transformed parameters {
  vector[n] mu = alpha
               + beta_CED * CED
               + beta_YN  * YN
               + beta_RL  * RL
               + beta_VA  * VA;
}

model {
  alpha    ~ normal(50, 20);
  beta_CED ~ normal(0, 1);
  beta_YN  ~ normal(0, 1);
  beta_RL  ~ normal(0, 1);
  beta_VA  ~ normal(0, 1);
  sigma    ~ normal(0, 10);

  VE ~ normal(mu, sigma);
}

generated quantities {
  vector[n] VE_rep;
  for (i in 1:n) {
    VE_rep[i] = normal_rng(mu[i], sigma);
  }
}"
) |>
  cmdstan_model()


fit <- df |>
  transmute(
    VE = SL.EMP.VULN.ZS |> scale() |> as.numeric(),
    CED = SE.COM.DURS |> scale() |> as.numeric(),
    YN = SL.UEM.NEET.ME.ZS |> scale() |> as.numeric(),
    RL = RL.EST |> scale() |> as.numeric(),
    VA = VA.EST |> scale() |> as.numeric()
  ) |>
  drop_na() |>
  compose_data() |>
  model$sample()
```

```{r}

fit$summary() |>
  filter(
    !stringr::str_detect(variable, "mu\\[") &
      !stringr::str_detect(variable, "VE_rep\\[")
  )

plot_param <- function(draws, param) {
  ggplot(draws, aes(x = .data[[param]])) +
    stat_function(
      fun = dnorm,
      args = list(mean = 0, sd = 1),
      linewidth = 1,
      color = "indianred3"
    ) +
    stat_halfeye(fill = "cornflowerblue", alpha = 0.7) +
    scale_x_continuous(limits = c(-2, 2)) +
    labs(title = param)
}

params <- c("beta_CED", "beta_YN", "beta_RL", "beta_VA")

draws <- fit |>
  spread_draws(alpha, beta_CED, beta_YN, beta_RL, beta_VA, sigma)

params |> purrr::walk(\(param) print(plot_param(draws, param)))

```