---
title: Suyash's project
engine: knitr
format:
  html:
    df-print: kable
    include-in-header:
      - text: |
          <script src="../js/dagitty.js"></script>
---

```{r project_config, echo=FALSE, warning=FALSE, message=FALSE}
variables <- c(
  GGR = 'NY.GDP.MKTP.KD.ZG',
  INF = 'FP.CPI.TOTL.ZG',
  PST = 'PV.PER.RNK',
  GEF = 'GE.PER.RNK',
  IMP = 'NE.IMP.GNFS.ZS'
)

histograms <- list(
  c(variables['INF'], variables['GGR'])
)

ALPHA_MEAN <- 50
ALPHA_SD <- 25
BETA_SD <- 10
SIGMA_SD <- 20

dag <- '
dag {
bb="0,0,1,1"
"GDP Growth" [outcome,pos="0.906,0.627"]
"Government Effectiveness" [adjusted,pos="0.387,0.179"]
"Import of Goods" [adjusted,pos="0.668,0.362"]
"Political Stability" [adjusted,pos="0.259,0.413"]
Inflation [exposure,pos="0.135,0.613"]
"Government Effectiveness" -> "Import of Goods"
"Government Effectiveness" -> "Political Stability"
"Import of Goods" -> "GDP Growth"
"Political Stability" -> "GDP Growth"
"Political Stability" -> Inflation
Inflation -> "GDP Growth"
}
'
```

{{< include _base.qmd >}}


# Formal Model
$$
\begin{align}
GGR_i &\sim \text{Normal}(\mu_i,\sigma) \\
\mu_i &= \alpha
+	\beta_{INF} INF_i
+	\beta_{PST} PST_i
+	\beta_{GEF} GEF_i
+	\beta_{IMP} IMP_i \\
\alpha &\sim \text{Normal}(`r ALPHA_MEAN`,`r ALPHA_SD`) \\
\beta_{INF} &\sim \text{Normal}(0,`r BETA_SD`) \\
\beta_{PST} &\sim \text{Normal}(0,`r BETA_SD`) \\
\beta_{GEF} &\sim \text{Normal}(0,`r BETA_SD`) \\
\beta_{IMP} &\sim \text{Normal}(0,`r BETA_SD`) \\
\sigma &\sim \text{Normal}(0,`r SIGMA_SD`), \quad \sigma > 0
\end{align}
$$

| Variable | Description | Unit |
|----------|-------------|------|
| $GGR_i$  | GDP Growth, annual | percentage |
| $INF_i$  | Inflation, annual, percent of GDP | standardized |
| $PST_i$  | Political Stability (percentile) | standardized |
| $GEF_i$  | Government Effectiveness (percentile) | standardized |
| $IMP_i$  | Import of Goods and Services (percent of GDP) | standardized |


# Base model

Just $GGR_i$ ~ $INF_i$

```{stan model1_stan, eval=FALSE, output.var="model1_stan"}
data {
  int<lower=1> n;
  vector[n] GGR;
  vector[n] INF;
}

parameters {
  real alpha;
  real beta_INF;
  real<lower=0> sigma;
}

transformed parameters {
  vector[n] mu = alpha + beta_INF * INF;
}

model {
  alpha    ~ normal(<<ALPHA_MEAN>>, <<ALPHA_SD>>);
  beta_INF ~ normal(0, <<BETA_SD>>);
  sigma    ~ normal(0, <<SIGMA_SD>>);

  GGR ~ normal(mu, sigma);
}

generated quantities {
  vector[n] GGR_rep;
  for (i in 1:n) {
    GGR_rep[i] = normal_rng(mu[i], sigma);
  }
}
```

```{r model1, message=FALSE, results="hide"}
model1 <- knitr::knit_code$get("model1_stan") |>
  paste(collapse = "\n") |>
  glue::glue(.open = "<<", .close = ">>") |>
  write_stan_file() |>
  cmdstan_model()

fit1 <- df |>
  transmute(
    GGR = .data[[variables['GGR']]],
    INF = .data[[variables['INF']]] |> scale() |> as.numeric(),
  ) |>
  drop_na() |>
  compose_data() |>
  model1$sample()
```

```{r}
fit1$summary() |>
  filter(
    !stringr::str_detect(variable, "mu\\[") &
      !stringr::str_detect(variable, "_rep\\[") &
      !stringr::str_detect(variable, "lp__")
  )

draws1 <- fit1 |>
  spread_draws(alpha, beta_INF, sigma)
plot_param(draws1, "beta_INF")
```

# Statistical model

Impact of Inflation on GDP Growth

Red is our prior distribution; blue is our posterior distribution.

```{stan model2_stan, eval=FALSE, output.var="model2_stan"}
data {
  int<lower=1> n;
  vector[n] GGR;
  vector[n] INF;
  vector[n] PST;
  vector[n] GEF;
  vector[n] IMP;
}

parameters {
  real alpha;
  real beta_INF;
  real beta_PST;
  real beta_GEF;
  real beta_IMP;
  real<lower=0> sigma;
}

transformed parameters {
  vector[n] mu = alpha
               + beta_INF * INF
               + beta_PST * PST
               + beta_GEF * GEF
               + beta_IMP * IMP;
}

model {
  alpha    ~ normal(<<ALPHA_MEAN>>, <<ALPHA_SD>>);
  beta_INF  ~ normal(0, <<BETA_SD>>);
  beta_PST  ~ normal(0, <<BETA_SD>>);
  beta_GEF  ~ normal(0, <<BETA_SD>>);
  beta_IMP  ~ normal(0, <<BETA_SD>>);
  sigma    ~ normal(0, <<SIGMA_SD>>);

  GGR ~ normal(mu, sigma);
}

generated quantities {
  vector[n] GGR_rep;
  for (i in 1:n) {
    GGR_rep[i] = normal_rng(mu[i], sigma);
  }
}
```

```{r model2, message=FALSE, results="hide"}
model2 <- knitr::knit_code$get("model2_stan") |>
  paste(collapse = "\n") |>
  glue::glue(.open = "<<", .close = ">>") |>
  write_stan_file() |>
  cmdstan_model()

fit2 <- df |>
  transmute(
    GGR = .data[[variables['GGR']]],
    INF = .data[[variables['INF']]] |> scale() |> as.numeric(),
    PST = .data[[variables['PST']]] |> scale() |> as.numeric(),
    GEF = .data[[variables['GEF']]] |> scale() |> as.numeric(),
    IMP = .data[[variables['IMP']]] |> scale() |> as.numeric()
  ) |>
  drop_na() |>
  compose_data() |>
  model2$sample()
```

```{r}
fit2$summary() |>
  filter(
    !stringr::str_detect(variable, "mu\\[") &
      !stringr::str_detect(variable, "_rep\\[") &
      !stringr::str_detect(variable, "lp__")
  )

params <- c("beta_INF", "beta_PST", "beta_GEF", "beta_IMP")
draws <- fit2 |>
  spread_draws(alpha, !!!syms(params), sigma)
params |> purrr::walk(\(param) print(plot_param(draws, param)))
```