---
title: Suyash's project
engine: knitr
format:
  html:
    df-print: kable
    include-in-header:
      - text: |
          <script src="../js/dagitty.js"></script>
---

```{r project_config, echo=FALSE, warning=FALSE, message=FALSE}
variables <- c(
  NMG = 'SM.POP.NETM.prop',
  YNE = 'SL.UEM.NEET.ME.ZS',
  AFR = 'SP.ADO.TFRT',
  YUE = 'SL.UEM.1524.ZS',
  GPC = 'NY.GDP.PCAP.PP.CD'
)

histograms <- list(
  c(variables['YNE'], variables['NMG'])
)

ALPHA_MEAN <- 0
ALPHA_SD <- 5
BETA_SD <- 10
SIGMA_SD <- 20

dag <- '
dag {
bb="0,0,1,1"
"Adolescent Fertility Rate" [adjusted,pos="0.471,0.457"]
"GDP (per capital)" [adjusted,pos="0.467,0.142"]
"Net Migration" [outcome,pos="0.927,0.623"]
"Youth NEETs" [exposure,pos="0.060,0.611"]
"Youth Unemployment" [adjusted,pos="0.469,0.321"]
"Adolescent Fertility Rate" -> "Net Migration"
"Adolescent Fertility Rate" -> "Youth NEETs"
"GDP (per capital)" -> "Net Migration"
"GDP (per capital)" -> "Youth NEETs"
"GDP (per capital)" -> "Youth Unemployment"
"Youth NEETs" -> "Net Migration"
"Youth Unemployment" -> "Net Migration"
"Youth Unemployment" -> "Youth NEETs"
}
'
```

{{< include _base.qmd >}}



# Formal Model
$$
\begin{align}
NMG_i &\sim \text{Normal}(\mu_i,\sigma) \\
\mu_i &= \alpha
+	\beta_{YNE} YNE_i
+	\beta_{AFR} AFR_i
+	\beta_{YUE} YUE_i
+	\beta_{GPC} GPC_i \\
\alpha &\sim \text{Normal}(`r ALPHA_MEAN`,`r ALPHA_SD`) \\
\beta_{YNE} &\sim \text{Normal}(0,`r BETA_SD`) \\
\beta_{AFR} &\sim \text{Normal}(0,`r BETA_SD`) \\
\beta_{YUE} &\sim \text{Normal}(0,`r BETA_SD`) \\
\beta_{GPC} &\sim \text{Normal}(0,`r BETA_SD`) \\
\sigma &\sim \text{Normal}(0,`r SIGMA_SD`), \quad \sigma > 0
\end{align}
$$

| Variable | Description | Unit |
|----------|-------------|------|
| $NMG_i$  | Net Migration | standardized |
| $YNE_i$  | Youth Not in Education, Employment or Training (%) | standardized |
| $AFR_i$  | Adolescent Fertility Rate (per 1000) | standardized |
| $YUE_i$  | Youth Unemployment Rate | standardized |
| $GPC_i$  | GDP per Capita (PPP) | standardized |


# Base model

Just $NMG_i$ ~ $YNE_i$

```{stan model1_stan, eval=FALSE, output.var="model1_stan"}
data {
  int<lower=1> n;
  vector[n] NMG;
  vector[n] YNE;
}

parameters {
  real alpha;
  real beta_YNE;
  real<lower=0> sigma;
}

transformed parameters {
  vector[n] mu = alpha + beta_YNE * YNE;
}

model {
  alpha    ~ normal(<<ALPHA_MEAN>>, <<ALPHA_SD>>);
  beta_YNE ~ normal(0, <<BETA_SD>>);
  sigma    ~ normal(0, <<SIGMA_SD>>);

  NMG ~ normal(mu, sigma);
}

generated quantities {
  vector[n] NMG_rep;
  for (i in 1:n) {
    NMG_rep[i] = normal_rng(mu[i], sigma);
  }
}
```

```{r model1, message=FALSE, results="hide"}
model1 <- knitr::knit_code$get("model1_stan") |>
  paste(collapse = "\n") |>
  glue::glue(.open = "<<", .close = ">>") |>
  write_stan_file() |>
  cmdstan_model()

fit1 <- df |>
  transmute(
    NMG = .data[[variables['NMG']]],
    YNE = .data[[variables['YNE']]] |> scale() |> as.numeric(),
  ) |>
  drop_na() |>
  compose_data() |>
  model1$sample()
```

```{r}
fit1$summary() |>
  filter(
    !stringr::str_detect(variable, "mu\\[") &
      !stringr::str_detect(variable, "_rep\\[") &
      !stringr::str_detect(variable, "lp__")
  )

draws1 <- fit1 |>
  spread_draws(alpha, beta_YNE, sigma)
plot_param(draws1, "beta_YNE")
```

# Statistical model

Impact of Youth NEET on Net Migration

Red is our prior distribution; blue is our posterior distribution.

```{stan model2_stan, eval=FALSE, output.var="model2_stan"}
data {
  int<lower=1> n;
  vector[n] NMG;
  vector[n] YNE;
  vector[n] AFR;
  vector[n] YUE;
  vector[n] GPC;
}

parameters {
  real alpha;
  real beta_YNE;
  real beta_AFR;
  real beta_YUE;
  real beta_GPC;
  real<lower=0> sigma;
}

transformed parameters {
  vector[n] mu = alpha
               + beta_YNE * YNE
               + beta_AFR * AFR
               + beta_YUE * YUE
               + beta_GPC * GPC;
}

model {
  alpha    ~ normal(<<ALPHA_MEAN>>, <<ALPHA_SD>>);
  beta_YNE  ~ normal(0, <<BETA_SD>>);
  beta_AFR  ~ normal(0, <<BETA_SD>>);
  beta_YUE  ~ normal(0, <<BETA_SD>>);
  beta_GPC  ~ normal(0, <<BETA_SD>>);
  sigma    ~ normal(0, <<SIGMA_SD>>);

  NMG ~ normal(mu, sigma);
}

generated quantities {
  vector[n] NMG_rep;
  for (i in 1:n) {
    NMG_rep[i] = normal_rng(mu[i], sigma);
  }
}
```

```{r model2, message=FALSE, results="hide"}
model2 <- knitr::knit_code$get("model2_stan") |>
  paste(collapse = "\n") |>
  glue::glue(.open = "<<", .close = ">>") |>
  write_stan_file() |>
  cmdstan_model()

fit2 <- df |>
  transmute(
    NMG = .data[[variables['NMG']]],
    YNE = .data[[variables['YNE']]] |> scale() |> as.numeric(),
    AFR = .data[[variables['AFR']]] |> scale() |> as.numeric(),
    YUE = .data[[variables['YUE']]] |> scale() |> as.numeric(),
    GPC = .data[[variables['GPC']]] |> scale() |> as.numeric()
  ) |>
  drop_na() |>
  compose_data() |>
  model2$sample()
```

```{r}
fit2$summary() |>
  filter(
    !stringr::str_detect(variable, "mu\\[") &
      !stringr::str_detect(variable, "_rep\\[") &
      !stringr::str_detect(variable, "lp__")
  )

params <- c("beta_YNE", "beta_AFR", "beta_YUE", "beta_GPC")
draws <- fit2 |>
  spread_draws(alpha, !!!syms(params), sigma)
params |> purrr::walk(\(param) print(plot_param(draws, param)))
```