---
title: Tsering's project
engine: knitr
format:
  html:
    df-print: kable
    include-in-header:
      - text: |
          <script src="../js/dagitty.js"></script>
---

<!-- TODO: come up with a better way to do string interpoplation to make the models easier to read -->
<!-- TODO: fix repetition of params in the spread draws -->

```{r project_config, echo=FALSE, warning=FALSE, message=FALSE}
variables <- c(
  'SL.EMP.VULN.ZS',
  'SL.EMP.TOTL.SP.FE.ZS',
  'NY.GDP.PCAP.PP.CD',
  'SL.UEM.NEET.ME.ZS',
  'SP.RUR.TOTL.ZS'
)

histograms <- list(
  c('SL.EMP.TOTL.SP.FE.ZS', 'SL.EMP.VULN.ZS')
)

ALPHA_MEAN <- 50
ALPHA_SD <- 25
BETA_SD <- 10
SIGMA_SD <- 20

dag <- '
dag {
"Rural Population" [latent,pos="0.493,0.601"]
"Vulnerable Employment" [outcome,pos="0.894,0.445"]
"Per Capita GDP" [pos="0.491,0.289"]
"Womens Employment" [exposure,pos="0.148,0.444"]
"Youth NEETs" [adjusted,pos="0.494,0.506"]
"Vulnerable Employment" -> "Rural Population"
"Per Capita GDP" -> "Vulnerable Employment"
"Womens Employment" -> "Rural Population"
"Womens Employment" -> "Vulnerable Employment"
"Womens Employment" -> "Per Capita GDP"
"Youth NEETs" -> "Vulnerable Employment"
"Youth NEETs" -> "Womens Employment"
}
'
```

{{< include _base.qmd >}}


# Formal Model
$$
\begin{align}
VE_i &\sim \text{Normal}(\mu_i,\sigma) \\
\mu_i &= \alpha
+	\beta_{WE} WE_i
+	\beta_{IN} IN_i
+	\beta_{YN} YN_i \\
\alpha &\sim \text{Normal}(`r ALPHA_MEAN`,`r ALPHA_SD`) \\
\beta_{WE} &\sim \text{Normal}(0,`r BETA_SD`) \\
\beta_{IN} &\sim \text{Normal}(0,`r BETA_SD`) \\
\beta_{YN} &\sim \text{Normal}(0,`r BETA_SD`) \\
\sigma &\sim \text{Normal}(0,`r SIGMA_SD`), \quad \sigma > 0
\end{align}
$$

| Variable | Description | Unit |
|----------|-------------|------|
| $VE_i$  | Vulnerable Employment | percentage |
| $WE_i$  | Women's Employment Rate | standardized |
| $IN_i$  | Per Capita GDP | standardized |
| $YN_i$  | Youth not in Education, Employment, or Training | standardized |


# Base Model

Just $VE_i$ ~ $WE_i$

```{r model1, message=FALSE, results="hide"}
model1 <- write_stan_file(
  paste0(
    "
data {
  int<lower=1> n;
  vector[n] VE;
  vector[n] WE;
}

parameters {
  real alpha;
  real beta_WE;
  real<lower=0> sigma;
}

transformed parameters {
  vector[n] mu = alpha + beta_WE * WE;
}

model {
  alpha    ~ normal(",
    ALPHA_MEAN,
    ", ",
    ALPHA_SD,
    ");
  beta_WE ~ normal(0, ",
    BETA_SD,
    ");
  sigma    ~ normal(0, ",
    SIGMA_SD,
    ");

  VE ~ normal(mu, sigma);
}

generated quantities {
  vector[n] VE_rep;
  for (i in 1:n) {
    VE_rep[i] = normal_rng(mu[i], sigma);
  }
}"
  )
) |>
  cmdstan_model()


fit1 <- df |>
  transmute(
    VE = SL.EMP.VULN.ZS,
    WE = SL.EMP.TOTL.SP.FE.ZS |> scale() |> as.numeric(),
  ) |>
  drop_na() |>
  compose_data() |>
  model1$sample()
```

```{r}
fit1$summary() |>
  filter(
    !stringr::str_detect(variable, "mu\\[") &
      !stringr::str_detect(variable, "VE_rep\\[") &
      !stringr::str_detect(variable, "lp__")
  )
```

# Statistical Model

Impact of Compulsory Education Duration on Vulnerable Employment rates

Red is our prior distribution; blue is our posterior distribution.
```{r model2, message=FALSE, results="hide"}
model2 <- write_stan_file(
  paste0(
    "
data {
  int<lower=1> n;
  vector[n] VE;
  vector[n] WE;
  vector[n] IN;
  vector[n] YN;
}

parameters {
  real alpha;
  real beta_WE;
  real beta_IN;
  real beta_YN;
  real<lower=0> sigma;
}

transformed parameters {
  vector[n] mu = alpha
               + beta_WE  * WE
               + beta_IN  * IN
               + beta_YN  * YN;
}

model {
  alpha    ~ normal(",
    ALPHA_MEAN,
    ", ",
    ALPHA_SD,
    ");
  beta_WE  ~ normal(0, ",
    BETA_SD,
    ");
  beta_IN  ~ normal(0, ",
    BETA_SD,
    ");
  beta_YN  ~ normal(0, ",
    BETA_SD,
    ");
  sigma    ~ normal(0, ",
    SIGMA_SD,
    ");

  VE ~ normal(mu, sigma);
}

generated quantities {
  vector[n] VE_rep;
  for (i in 1:n) {
    VE_rep[i] = normal_rng(mu[i], sigma);
  }
}"
  )
) |>
  cmdstan_model()


fit2 <- df |>
  transmute(
    VE = SL.EMP.VULN.ZS,
    WE = SL.EMP.TOTL.SP.FE.ZS |> scale() |> as.numeric(),
    IN = NY.GDP.PCAP.PP.CD |> scale() |> as.numeric(),
    YN = SL.UEM.NEET.ME.ZS |> scale() |> as.numeric()
  ) |>
  drop_na() |>
  compose_data() |>
  model2$sample()
```

```{r}

fit2$summary() |>
  filter(
    !stringr::str_detect(variable, "mu\\[") &
      !stringr::str_detect(variable, "VE_rep\\[") &
      !stringr::str_detect(variable, "lp__")
  )

params <- c("beta_WE", "beta_IN", "beta_YN")

draws <- fit2 |>
  spread_draws(alpha, beta_WE, beta_IN, beta_YN, sigma)

params |> purrr::walk(\(param) print(plot_param(draws, param)))

```