---
format:
  revealjs:
    include-in-header:
      - text: |
          <style>
            .workflow {
              display: flex;
              gap: 14px;
              align-items: stretch;
              margin: 30px 0;
            }
            .workflow .step {
              flex: 1 1 0;
              border: 2px solid #16314d;
              border-radius: 10px;
              padding: 18px 12px;
              text-align: center;
              font-weight: 700;
              font-size: 0.8em;
              background: white;
            }
            .workflow .step small {
              display: block;
              font-weight: 400;
              font-size: 0.7em;
              margin-top: 8px;
              color: #16314d;
            }
            .workflow .arrow {
              align-self: center;
              font-size: 1.4em;
              color: #999;
            }
            .workflow .step.science {
              background: #e8f3ff;
            }
          </style>
---

```{r setup, include=FALSE}
library(tidyverse)
library(cmdstanr)
library(tidybayes)
source("r/rlib.R")

knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 12,
  fig.height = 6.5,
  dpi = 150,
  dev.args = list(bg = "transparent")
)

BLUE <- "#1b91ff"
RED <- "#ff2c2d"
GRAY <- "#999999"
INK <- "#16314d"

theme_set(
  theme_minimal(base_size = 22) +
    theme(
      panel.grid.minor = element_blank(),
      plot.title = element_text(face = "bold", color = INK)
    )
)

fmt <- function(x, d = 1) formatC(x, format = "f", digits = d, big.mark = ",")

# Fit a Stan model quietly
fit_stan <- function(file, data) {
  cmdstan_model(file, dir = ".stancache")$sample(
    data = data,
    seed = 1,
    refresh = 0,
    show_messages = FALSE,
    show_exceptions = FALSE
  )
}

# Posterior of (male - female) from a model with alpha[sex]
sex_difference <- function(fit) {
  fit$draws("alpha", format = "df") |>
    as_tibble() |>
    transmute(difference = `alpha[2]` - `alpha[1]`)
}

# Plot posteriors of one quantity under several models, one row per model
plot_posteriors <- function(draws, x, xlab, colors = c(RED, BLUE)) {
  ggplot(draws, aes(x = {{ x }}, y = model, fill = model)) +
    geom_vline(xintercept = 0, color = GRAY, linetype = "dashed") +
    stat_halfeye(.width = 0.89, point_size = 4, slab_alpha = 0.8, normalize = "groups") +
    scale_fill_manual(values = colors, guide = "none") +
    labs(x = xlab, y = NULL)
}

# !Kung San adults (Howell), from Week 2
howell <- read_csv("datasets/kungsan.csv") |>
  select(-1) |>
  filter(age >= 18) |>
  mutate(
    sex = factor(if_else(male == 1, "male", "female"), levels = c("female", "male")),
    height_std = (height - mean(height)) / sd(height)
  )

# Ames, Iowa house sales, from Week 2
houses <- read_csv("datasets/ames.csv") |>
  mutate(
    z_price = (price - mean(price)) / sd(price),
    z_bedrooms = (bedrooms - mean(bedrooms)) / sd(bedrooms),
    z_area = (area_m2 - mean(area_m2)) / sd(area_m2)
  )

# World Bank 2022: fertility, women's college enrollment, infant mortality
countries <- read_csv(
  "datasets/wb2022.csv",
  na = c("", ".."),
  col_types = cols(.default = "c")
) |>
  select(code = `Country Code`, series = `Series Code`, value = `2022 [YR2022]`) |>
  filter(series %in% c("SP.DYN.TFRT.IN", "SE.TER.ENRR.FE", "SP.DYN.IMRT.IN")) |>
  distinct(code, series, .keep_all = TRUE) |>
  mutate(value = as.numeric(value)) |>
  pivot_wider(names_from = series, values_from = value) |>
  transmute(
    code,
    fertility_rate = SP.DYN.TFRT.IN,
    college = SE.TER.ENRR.FE,
    mortality = SP.DYN.IMRT.IN
  ) |>
  drop_na() |>
  mutate(
    women_in_college = (college - mean(college)) / sd(college),
    infant_mortality = (mortality - mean(mortality)) / sd(mortality)
  )
```



# Data Lunch: Week 3 {.center}

Causal models


# The doctor paradox

## {auto-animate=true}
Two surgeons, one normal and one famous, have different success rates for a procedure. How should we interpret this?

::: {data-id="table"}
<table>
  <thead>
    <tr>
      <td data-id="leftmargin1"></td>
      <td data-id="rightmargin1">Success</td>
    </tr>
  </thead>
  <tbody>
    <tr>
      <td data-id="leftmargin2">Normal&nbsp;Surgeon</td>
      <td data-id="rightmargin2">[89.6%]{.green}</td>
    </tr>
    <tr>
      <td data-id="leftmargin3">Famous&nbsp;Surgeon</td>
      <td data-id="rightmargin3">[74.9%]{.red  }</td>
    </tr>
  </tbody>
</table>
:::

## {auto-animate=true}
How about now?

::: {data-id="table"}
<table>
  <thead>
    <tr>
      <td data-id="leftmargin1"></td>
      <td>Standard</td>
      <td>Complex</td>
      <td>Critical</td>
      <td  data-id="rightmargin1" style="border-left: 1px solid black;">Total</td>
    </tr>
  </thead>
  <tbody>
    <tr>
      <td data-id="leftmargin2">Normal&nbsp;Surgeon</td>
      <td>[97.3%]{.red}</td>
      <td>[66.0%]{.red  }</td>
      <td>[38.0%]{.red  }</td>
      <td data-id="rightmargin2" style="border-left: 1px solid black;">[89.6%]{.green}</td>
    </tr>
    <tr>
      <td data-id="leftmargin3">Famous&nbsp;Surgeon</td>
      <td>[99.7%]{.green}</td>
      <td>[75.0%]{.green}</td>
      <td>[50.0%]{.green}</td>
      <td data-id="rightmargin3" style="border-left: 1px solid black;">[74.9%]{.red  }</td>
    </tr>
  </tbody>
</table>
:::


## What changed?

::: {.columns}
::: {.column width="50%"}
```{r surgeon-dag, fig.width = 7, fig.height = 5.5}
dag_plot(
  c("C -> S", "C -> O", "S -> O"),
  pos = list(C = c(1, 1.2), S = c(0, 0), O = c(2, 0)),
  labels = c(C = "Case difficulty", S = "Surgeon", O = "Outcome"),
  above = "C",
  exposure = "S",
  outcome = "O"
)
```
:::

::: {.column width="50%"}
- The famous surgeon gets the **hardest cases**.
- Hard cases fail more often, whoever operates.
- Case difficulty **interferes**: it affects *who operates* and *whether the operation
  succeeds*.
- Adding one variable to the comparison **reversed** the finding.
:::
:::


## Which number is right?

- Both tables are calculated correctly from the same records.
- They answer **different questions**:
  - "Whose patients did better last year?" → [the normal surgeon]{.green}
  - "For *my* operation, who gives me the better chance?" → [the famous surgeon]{.green}
- The data alone can't tell us which comparison to make.
- We need a story about **how the data was made**. That story is a **causal model**.


# Goals this week

- Part 1: Concepts
  - DAGs: drawing what causes what
  - What "cause" means in statistics
  - The workflow: estimand → generative model → statistical model → data → estimate
  - Worked example: does sex affect weight?
- Part 2: Applied
  - Designing DAGs for your own questions


# What is a DAG?

::: {.columns}
::: {.column width="55%"}
- **D**irected: arrows point one way, from cause to effect.
- **A**cyclic: no loops. Nothing causes itself.
- **G**raph: variables (circles) joined by arrows.
- An arrow means "**directly causes**".
- A **missing** arrow is a claim too: "no direct effect".
- A DAG shows *what depends on what*, not *how much*.
- It's a picture of our **assumptions**, drawn from what we know about the world, not a
  result we get from data.
:::

::: {.column width="45%"}
```{r dag-anatomy, fig.width = 6, fig.height = 4.5}
dag_plot(
  c("C -> S", "C -> O", "S -> O"),
  pos = list(C = c(1, 1.2), S = c(0, 0), O = c(2, 0)),
  labels = c(C = "Case difficulty", S = "Surgeon", O = "Outcome"),
  above = "C",
  exposure = "S",
  outcome = "O"
)
```

[**Exposure**]{.blue}: the cause we're asking about. **Outcome**: the effect we
care about.
:::
:::



# What is a "cause"?

- In statistics, X **causes** Y if *intervening* to change X would change Y, on average.
  - "If this patient were switched to the famous surgeon, would their chance of
    success change?"
- Causes are **probabilistic**: smoking causes lung cancer, even though most smokers
  never get it.
- Causes are **many**: almost every outcome has lots of causes acting at once.
- A causal claim is always a **simplification** of a messy world. In social science we
  usually think of causes as places where we could *intervene*: a policy, a treatment,
  a decision.


## Seeing vs. doing

::: {.columns}
::: {.column}
::: {.callout-note .nonincremental title="Seeing"}
P( success | famous surgeon )

Among patients who *happened* to get the famous surgeon, how many recovered?

Includes all his hard cases.
:::
:::

::: {.column}
::: {.callout-tip .nonincremental title="Doing"}
P( success | do(famous surgeon) )

If we *assigned* patients to the famous surgeon, how many would recover?

This is the causal question.
:::
:::
:::

- Data gives us **seeing**. Most of our questions are about **doing**.
- A causal model is the bridge between them.


# The workflow

```{=html}
<div class="workflow">
  <div class="step science">Estimand<small>What exactly do we want to know?</small></div>
  <div class="arrow">→</div>
  <div class="step science">Generative model<small>How do we think the data was made?</small></div>
  <div class="arrow">→</div>
  <div class="step">Statistical model<small>What should we fit?</small></div>
  <div class="arrow">→</div>
  <div class="step">Data<small>What did we observe?</small></div>
  <div class="arrow">→</div>
  <div class="step">Estimate<small>Which values are most plausible?</small></div>
</div>
```

- **Estimand**: the question, stated precisely enough that a number could answer it.
- **Generative model**: our causal story (a DAG, plus a simulation of it).
- **Statistical model**: derived *from* the generative model, not picked from a menu.
- **Estimate**: the posterior. *Given what we've seen, which values of the estimand are
  most plausible?*
- The [shaded]{.blue} steps are science. They come first, before we touch the data.


# Worked example: does sex affect weight?

```{r howell-scatter, fig.height = 5.5}
ggplot(howell, aes(height, weight, color = sex)) +
  geom_point(alpha = 0.7, size = 2.5) +
  scale_color_manual(values = c(female = RED, male = BLUE), name = NULL) +
  labs(x = "Height (cm)", y = "Weight (kg)") +
  theme(legend.position = "top")
```

- The !Kung San adults from Week 2. Men are taller *and* heavier.


## Step 1: The estimand

- "Does sex affect weight?" is too vague to answer.
- More precisely:

::: {.callout-tip .nonincremental title="Estimand"}
How much heavier, on average, would an adult be if they were male rather than female?
:::

- The answer is a number, in kg: the **effect of sex on weight**.


## Step 2: The generative model

::: {.columns}
::: {.column width="50%"}
```{r howell-dag, fig.width = 7, fig.height = 4.2}
howell_edges <- c("S -> H", "S -> W", "H -> W", "A -> H", "A -> W")
howell_pos <- list(S = c(0, 1), H = c(1, 0), W = c(2, 1), A = c(3, 0))
howell_labels <- c(S = "Sex", H = "Height", W = "Weight", A = "Age")

dag_plot(
  howell_edges,
  howell_pos,
  labels = howell_labels,
  above = c("S", "W"),
  exposure = "S",
  outcome = "W"
)
```
:::

::: {.column width="50%"}
- Sex affects height (S → H).
- Height affects weight (H → W).
- Sex might also affect weight directly, e.g. through muscle mass (S → W).
- Age affects height and weight (A → H, A → W).
- Nothing in this picture causes sex.
:::
:::


## Step 2: Simulate the world we drew

::: {.columns}
::: {.column width="50%"}
```{r simulate, echo = TRUE}
simulate_people <- function(n) {
  tibble(
    sex = sample(c("female", "male"), n, TRUE),
    male = sex == "male",
    # S -> H: men 10 cm taller
    height = rnorm(n, 150 + 10 * male, 6),
    # H -> W: 0.6 kg per cm
    # S -> W: men 2 kg heavier at the
    #         same height
    weight = rnorm(
      n,
      -45 + 0.6 * height + 2 * male,
      4
    )
  )
}
```
:::

::: {.column width="50%"}
```{r sim-plot, fig.width = 7, fig.height = 5}
set.seed(4)
sim <- simulate_people(1000) |>
  mutate(
    sex = factor(sex, levels = c("female", "male")),
    height_std = (height - mean(height)) / sd(height)
  )

ggplot(sim, aes(height, weight, color = sex)) +
  geom_point(alpha = 0.5, size = 2) +
  scale_color_manual(values = c(female = RED, male = BLUE), name = NULL) +
  labs(x = "Height (cm)", y = "Weight (kg)") +
  theme(legend.position = "top")
```
:::
:::

- In this made-up world we **know** the answers:
  - Through the direct path alone: **2 kg**.
  - Through both paths: 2 + 0.6 × 10 = **8 kg**.


## Generative Model Demo {.no-title}

```{shinylive-r}
#| standalone: true
{{< include shiny/regression.R >}}
```


## Step 3: Two statistical models

::: {.columns}
::: {.column}
**Model A**: weight ~ sex

$$
\begin{align}
W_i &\sim \text{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha_{S[i]} \\
\alpha_j &\sim \text{Normal}(50, 10) \\
\sigma &\sim \text{Normal}^+(0, 10)
\end{align}
$$
:::

::: {.column}
**Model B**: weight ~ sex + height

$$
\begin{align}
W_i &\sim \text{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha_{S[i]} + \beta H_i \\
\alpha_j &\sim \text{Normal}(50, 10) \\
\beta &\sim \text{Normal}(0, 2) \\
\sigma &\sim \text{Normal}^+(0, 10)
\end{align}
$$
:::
:::

- In both, our estimate is the difference: α~male~ − α~female~.


## Step 3: Test the models on simulated data

```{r sim-fits, results = "hide"}
sim_a <- fit_stan("models/weight2.stan", compose_data(select(sim, weight, sex)))
sim_b <- fit_stan(
  "models/weight4.stan",
  compose_data(select(sim, weight, height_std, sex))
)

sim_draws <- bind_rows(
  sex_difference(sim_a) |> mutate(model = "A: weight ~ sex"),
  sex_difference(sim_b) |> mutate(model = "B: weight ~ sex + height")
)
```

```{r sim-posteriors, fig.height = 5}
plot_posteriors(sim_draws, difference, "Men − women (kg)") +
  geom_vline(xintercept = c(2, 8), color = INK, linewidth = 1) +
  annotate("text", x = 8, y = 2.6, label = "both paths: 8 kg", hjust = 1.05, size = 7) +
  annotate("text", x = 2, y = 2.6, label = "direct: 2 kg", hjust = -0.05, size = 7)
```

- Model A lands near **8 kg**. Model B lands near **2 kg**.
- Each model correctly recovers a *different* true number. Which one we want depends
  on the estimand.


## Steps 4 & 5: Real data, real estimates

```{r howell-fits, results = "hide"}
fit_a <- fit_stan("models/weight2.stan", compose_data(select(howell, weight, sex)))
fit_b <- fit_stan(
  "models/weight4.stan",
  compose_data(select(howell, weight, height_std, sex))
)

howell_draws <- bind_rows(
  sex_difference(fit_a) |> mutate(model = "A: weight ~ sex"),
  sex_difference(fit_b) |> mutate(model = "B: weight ~ sex + height")
)
howell_summary <- howell_draws |>
  group_by(model) |>
  median_qi(difference, .width = 0.89)
```

```{r howell-posteriors, fig.height = 5}
plot_posteriors(howell_draws, difference, "Men − women (kg)")
```

- Model A: men are about **`r fmt(howell_summary$difference[1])` kg** heavier.
- Model B: at the same height, men are about **`r fmt(howell_summary$difference[2])` kg**
  heavier. In other words, no difference.


## Effects flow through paths

::: {.columns}
::: {.column}
```{r flow-total, fig.width = 7, fig.height = 4}
dag_plot(
  howell_edges,
  howell_pos,
  labels = howell_labels,
  above = c("S", "W"),
  highlight = c("S -> W", "S -> H", "H -> W"),
  exposure = "S",
  outcome = "W"
)
```

**Model A** lets the effect flow through **both** paths: the **total effect**.
:::

::: {.column}
```{r flow-direct, fig.width = 7, fig.height = 4}
dag_plot(
  howell_edges,
  howell_pos,
  labels = howell_labels,
  above = c("S", "W"),
  highlight = "S -> W",
  exposure = "S",
  outcome = "W"
)
```

**Model B** holds height fixed, which **blocks** the path through height: only the
**direct effect** is left.
:::
:::

- Sex affects weight almost entirely *through height*.
- Our estimand asked for the total effect, so **Model A** answers it. Adding height
  would hide the very effect we're asking about.


## What about age?

```{r age-check}
direct_with_age <- coef(lm(weight ~ sex + height + age, data = howell))[["sexmale"]]
direct_without_age <- coef(lm(weight ~ sex + height, data = howell))[["sexmale"]]
```

::: {.columns}
::: {.column width="50%"}
```{r age-dag, fig.width = 7, fig.height = 4.2}
dag_plot(
  howell_edges,
  howell_pos,
  labels = howell_labels,
  above = c("S", "W"),
  highlight = c("A -> H", "A -> W"),
  exposure = "S",
  outcome = "W"
)
```
:::

::: {.column width="50%"}
- Age is in the DAG, but nothing points from age **into sex**, so age opens no extra
  path from sex to weight.
- So for this estimand we can leave age out.
- Quick check: adding age to the model shifts the sex difference by only
  `r fmt(abs(direct_with_age - direct_without_age), 2)` kg.
:::
:::

- **Not every variable in the DAG belongs in the model.** The DAG tells us which ones do.


# Remember the bedrooms?

::: {.columns}
::: {.column width="45%"}
```{r house-dag, fig.width = 6, fig.height = 5}
house_edges <- c("A -> B", "A -> P", "B -> P")
house_pos <- list(A = c(1, 1.2), B = c(0, 0), P = c(2, 0))
house_labels <- c(A = "Living area", B = "Bedrooms", P = "Price")

dag_plot(
  house_edges,
  house_pos,
  labels = house_labels,
  above = "A",
  highlight = c("A -> B", "A -> P"),
  exposure = "B",
  outcome = "P"
)
```
:::

::: {.column width="55%"}
- Last week: more bedrooms go with a *higher* price, but among houses of similar size,
  more bedrooms go with a *lower* price.
- Bigger houses have more bedrooms **and** sell for more.
- So the bedroom–price correlation also flows along [B ← A → P]{.red}: a path that
  isn't an effect of bedrooms at all.
:::
:::


## Bedrooms: two models

```{r house-fits, results = "hide"}
house_fit <- function(predictors) {
  fit_stan(
    "models/linear.stan",
    list(
      n = nrow(houses),
      k = length(predictors),
      X = as.matrix(houses[predictors]),
      y = houses$z_price
    )
  )
}

house_1 <- house_fit("z_bedrooms")
house_2 <- house_fit(c("z_bedrooms", "z_area"))

house_draws <- bind_rows(
  house_1$draws("beta[1]", format = "df") |> mutate(model = "price ~ bedrooms"),
  house_2$draws("beta[1]", format = "df") |> mutate(model = "price ~ bedrooms + area")
) |>
  as_tibble() |>
  rename(beta_bedrooms = `beta[1]`)
```

```{r house-posteriors, fig.height = 4.5}
plot_posteriors(house_draws, beta_bedrooms, "β for bedrooms (SD of price per SD of bedrooms)")
```

- Holding area fixed **reverses the sign**. An extra bedroom in the same space means
  smaller rooms.
- Here we *should* hold area fixed: the path through area isn't the effect we want.
  With height, we *shouldn't*. Same move, opposite advice. **The DAG decides.**


# Part 1: The punchline

- The variables we put in a model can change, or even reverse, the answer.
- That's because effects **flow along paths** through the DAG, and each variable we add
  can block some of them.
- So choosing variables is a **causal** decision, not a statistical one.
- The DAG comes first. The model follows from it.


# Part 2: Designing DAGs {.center}

The science has to come first.


## The data can't draw the DAG for you

- Many different DAGs fit the same data equally well.
- The arrows come from what you know: theory, earlier research, fieldwork, experience,
  common sense.
- If this feels **arbitrary**: good. You've noticed that every conclusion rests on
  assumptions.
- The job isn't to avoid assumptions. It's to make them **explicit**, so other people can
  argue with them.


## A recipe

1. **Name the estimand**: "the effect of X on Y".
2. List the things that **cause Y**.
3. List the things that **cause X**.
4. Draw an arrow only where you believe there's a **direct** cause. Every missing arrow is
   a claim.
5. Include things you **can't measure**, too (dashed circles).
6. Trace every **path** between X and Y. Which carry the effect you want? Which don't?


## Worked example: education and birth rates

::: {.columns}
::: {.column width="55%"}
```{r fertility-scatter, fig.width = 7, fig.height = 5.5}
ggplot(countries, aes(college, fertility_rate)) +
  geom_point(color = BLUE, alpha = 0.7, size = 3) +
  geom_smooth(method = "lm", se = FALSE, color = INK) +
  labs(
    x = "Women's college enrollment (% gross)",
    y = "Births per woman"
  )
```
:::

::: {.column width="45%"}
- World Bank, 2022: `r nrow(countries)` countries.
- **Estimand**: the effect of women's college enrollment on the birth rate.
- Countries where more women go to college have fewer births per woman.
- But what else is going on?
:::
:::


## Two defensible DAGs

```{r fertility-dags}
fert_labels <- c(
  C = "Women in college",
  M = "Infant mortality",
  F = "Births per woman",
  D = "Development"
)
```

::: {.columns}
::: {.column}
**DAG 1: development first**

```{r fertility-dag1, fig.width = 7, fig.height = 4.5}
dag_plot(
  c("D -> C", "D -> M", "C -> F", "M -> F"),
  pos = list(D = c(0.4, 1.4), C = c(0, 0), M = c(2.6, 1.4), F = c(3, 0)),
  labels = fert_labels,
  above = c("D", "M"),
  highlight = c("D -> C", "D -> M", "M -> F"),
  exposure = "C",
  outcome = "F",
  unobserved = "D"
)
```

Richer countries have more women in college *and* fewer infant deaths. The red path
isn't an effect of education. **Hold infant mortality fixed.**
:::

::: {.column}
**DAG 2: education first**

```{r fertility-dag2, fig.width = 7, fig.height = 4.5}
dag_plot(
  c("C -> M", "C -> F", "M -> F"),
  pos = list(C = c(0, 0), M = c(1.5, 1.2), F = c(3, 0)),
  labels = fert_labels,
  above = "M",
  highlight = c("C -> M", "M -> F"),
  exposure = "C",
  outcome = "F"
)
```

Educated mothers have healthier babies, so families need fewer births. The red path
*is* part of education's effect. **Don't hold infant mortality fixed.**
:::
:::


## Same data, different science, different answer

```{r fertility-fits, results = "hide"}
fert_1 <- fit_stan(
  "models/fertility1a.stan",
  compose_data(select(countries, fertility_rate, women_in_college))
)
fert_2 <- fit_stan(
  "models/fertility2.stan",
  compose_data(select(countries, fertility_rate, women_in_college, infant_mortality))
)

fert_draws <- bind_rows(
  fert_1$draws("beta", format = "df") |>
    rename(beta_college = beta) |>
    mutate(model = "DAG 2: births ~ college"),
  fert_2$draws("beta_college", format = "df") |>
    mutate(model = "DAG 1: births ~ college + mortality")
) |>
  as_tibble()

fert_summary <- fert_draws |>
  group_by(model) |>
  median_qi(beta_college, .width = 0.89)
```

```{r fertility-posteriors, fig.height = 4.5}
plot_posteriors(fert_draws, beta_college, "β for college (births per woman, per SD of enrollment)")
```

- If you believe **DAG 1**, education has a small effect
  (`r fmt(fert_summary$beta_college[fert_summary$model == "DAG 1: births ~ college + mortality"], 2)`).
- If you believe **DAG 2**, a large one
  (`r fmt(fert_summary$beta_college[fert_summary$model == "DAG 2: births ~ college"], 2)`).
- Which do you believe, and **why**? What would convince someone who believes the other?


## Your turn: draw your project's DAG

::: {.columns}
::: {.column width="55%"}
1. What are your **X** and **Y**?
2. Draw it at [dagitty.net](https://dagitty.net/dags.html) (next slide).
3. Which arrow are you **least** sure about?
4. What would a skeptic **add**?
5. Is anything important **unmeasured**?
6. Trace the paths from X to Y. Which should your model **block**?
:::

::: {.column width="45%"}
Your project file already has a `dag`. It's the same drawing, written as text:

```
dag {
  College -> Fertility
  Mortality -> Fertility
  Development -> College
  Development -> Mortality
}
```
:::
:::


## DAGitty {.no-title}

```{=html}
<iframe
  data-external="1"
  style="width:100%; aspect-ratio: 2 / 1; border:1px solid black;"
  src="https://dagitty.net/dags.html"
></iframe>
```


# Summary

- A **DAG** is a drawing of our causal assumptions: arrows for direct causes.
- **Cause** means "changing X would change Y". Data shows us *seeing*; we usually want
  *doing*.
- **Workflow**: estimand → generative model → statistical model → data → estimate.
- The variables we include can change, even reverse, our answers, because effects flow
  along different paths.
- The science comes first. Next week: the building blocks of DAGs, and rules for which
  paths to block.
