---
format:
  revealjs: default
---

```{r setup, include=FALSE}
library(tidyverse)
source("r/rlib.R")

knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 12,
  fig.height = 6.5,
  dpi = 150,
  dev.args = list(bg = "transparent")
)

BLUE <- "#1b91ff"
RED <- "#ff2c2d"
GRAY <- "#999999"
INK <- "#16314d"

theme_set(
  theme_minimal(base_size = 22) +
    theme(
      panel.grid.minor = element_blank(),
      plot.title = element_text(face = "bold", color = INK),
      legend.position = "top"
    )
)

fmt <- function(x, d = 1) formatC(x, format = "f", digits = d, big.mark = ",")
pct <- function(x) paste0(round(100 * x), "%")
```


# Data Lunch: Week 5 {.center}

Reading a finding, asking a question


# Should we act on this?

<!-- TODO: a real headline (ideally Nepal or South Asia) making a confident causal claim -->

> _"[Headline goes here]"_

- Hands up: would you fund a program based on this?
- Keep your answer. We'll vote again at the end of Part 1.


# Last four weeks → this week

- **Week 1**: P(A|B) ≠ P(B|A). Surprising findings are usually false.
- **Week 2**: Describe numbers by distributions. Coefficients have units.
- **Week 3**: Seeing ≠ doing. The DAG decides which number answers your question.
- **Week 4**: Forks, pipes, colliders. Holding the wrong thing fixed creates paradoxes.
- This week: put it together, from **both sides** of an analysis.


# Goals this week

- Part 1: Evaluating someone else's finding
  - Seven questions to ask of any claim
  - Worked case: bicycles for girls in Bihar
- Part 2: Asking your own question
  - From a decision to an estimand
  - The target trial: the experiment you wish you could run
- Part 3: Applied
  - Critique a study, design an estimand, revisit your project


# Part 1: Evaluating a finding {.center}

Seven questions


## Seven questions

::: {.nonincremental}
1. **What exactly is the claim?**
2. **What did they actually measure?**
3. **How was the data made?**
4. **What's in their DAG, and what's in their model?**
5. **Who is it about?**
6. **How big, and how sure?**
7. **How plausible was it before?**
:::

::: {.fragment}
Every one of these comes from a week you've already done, except one or two new pieces.
:::


# 1. What exactly is the claim?

::: {.columns}
::: {.column width="33%"}
::: {.callout-note .nonincremental title="Descriptive"}
_How much? How many?_

"38% of Kathmandu voters support the mayor."
:::
:::

::: {.column width="33%"}
::: {.callout-tip .nonincremental title="Predictive"}
_What will happen?_

"The mayor will probably win."
:::
:::

::: {.column width="33%"}
::: {.callout-important .nonincremental title="Causal"}
_What if we intervened?_

"Free school meals raise attendance."
:::
:::
:::

- Each kind needs different evidence. Causal claims need the most.
- Watch the verbs: _boosts, drives, leads to, protects against_ are causal claims, even
  when the study was only able to support a descriptive one.
- Exercise: rewrite the headline as an **estimand**. Is it seeing or doing? _(Week 3)_

<!-- TODO: 3-4 real headlines; students classify each and spot causal verbs on descriptive evidence -->


# 2. What did they actually measure?

- The **concept** in the headline vs. the **variable** in the data.
  - "Education" → years of schooling? enrollment? test scores?
  - "Women's college enrollment (% gross)": can exceed 100%. Why?
- **Proxies** are descendants: they carry the concept only partly. _(Week 4)_
- **Missing data**: who didn't answer, and why? Missingness is often selection on a collider.
- Measured once, or tracked over time? Self-reported, or observed?

<!-- TODO: example of a measure that drifts from its concept (e.g. "poverty" as income vs. consumption) -->


# 3. How was the data made?

::: {.columns}
::: {.column}
```{r obs-dag, fig.width = 7, fig.height = 4.5}
dag_plot(
  c("U -> X", "U -> Y", "X -> Y"),
  list(X = c(0, 0), U = c(1, 1.1), Y = c(2, 0)),
  labels = c(X = "Program", U = "Motivation, wealth, ...", Y = "Outcome"),
  above = "U",
  highlight = c("U -> X", "U -> Y"),
  exposure = "X",
  outcome = "Y",
  unobserved = "U"
)
```

**Observational**: people chose X. Everything that drove that choice is a possible
back door.
:::

::: {.column}
::: {.fragment}
```{r rct-dag, fig.width = 7, fig.height = 4.5}
dag_plot(
  c("R -> X", "U -> Y", "X -> Y"),
  list(X = c(0, 0), U = c(1, 1.1), Y = c(2, 0), R = c(-0.2, 1.1)),
  labels = c(R = "Coin flip", X = "Program", U = "Motivation, wealth, ...", Y = "Outcome"),
  above = c("U", "R"),
  exposure = "X",
  outcome = "Y",
  unobserved = "U"
)
```

**Randomized**: a coin flip chose X. Every arrow into X is **deleted**. This is
`do(X)` done physically.
:::
:::
:::


## Between the extremes

- **Randomized experiments**: strongest for "does it work _here, for these people_?"
- **Natural experiments**: something _like_ a coin flip (a lottery, a cutoff, a border,
  a policy rollout date).
- **Before / after, with a comparison group**: works if the groups would have moved
  together without the program.
- **Cross-sectional comparison**: one snapshot, many back doors. Everything in Week 4
  applies.

::: {.fragment}
Ask: **"What decided who got X?"** If the answer is "people chose," start drawing back
doors.
:::

<!-- TODO: regression to the mean slide here, or in Week 2: "we intervened on the worst
schools and they improved" (before/after with no comparison group) -->


# 4. What's in their DAG, and what's in their model?

Most reports never draw a DAG. Draw it for them.

::: {.nonincremental}
- Which **forks** did they control for? Which did they miss?
- Did they control for a **mediator**, hiding the effect? ("after adjusting for...")
- Did they select on a **collider**? ("among patients," "among graduates,"
  "among funded firms")
:::

::: {.fragment}
The phrase **"after controlling for..."** is the most important sentence in the
methods section. Check every variable in that list against your DAG. _(Week 4)_
:::


# 5. Who is it about?

- **Sample vs. population**: who was studied, and who are you making a decision for?
- **Would it work here?** Different place, time, prices, institutions, baseline rates.
- **Average vs. everyone**: an average effect can hide people it hurts.

::: {.fragment .callout-important .nonincremental title="Ecological fallacy"}
Countries with more X have more Y does **not** mean people with more X have more Y.

Your projects use country-level data. What can they say about individual women,
households, students?
:::

<!-- TODO: classic ecological example (e.g. Robinson 1950, literacy and immigration) -->


# 6. How big, and how sure?

::: {.columns}
::: {.column}
**How big**

- Translate into units that matter for the decision: rupees, students, lives per
  10,000.
- **Relative vs. absolute**: "doubles your risk" from 1 in 10,000 to 2 in 10,000 is
  one extra case per 10,000 people.
- Compare to the cost, and to the alternatives.
:::

::: {.column}
**How sure**

- Look at the **range** of plausible values, not just the middle.
- Does the decision change anywhere in that range?
- If the range runs from "large benefit" to "slight harm," you don't know yet. That is
  a finding too.
:::
:::


## Reading a table you're handed

Most reports you'll read aren't Bayesian. You can still read them.

<!-- TODO: replace with a real (or realistic) regression table from a report -->

| | Estimate | Std. error | |
|---|---:|---:|---|
| Program | 0.42 | 0.15 | \*\* |
| Household income (log) | 1.10 | 0.30 | \*\*\* |
| Distance to school (km) | −0.08 | 0.06 | |
: {.nonincremental}

- **Rough plausible range**: estimate ± 2 × std. error.
  - Program: `r fmt(0.42 - 2 * 0.15, 2)` to `r fmt(0.42 + 2 * 0.15, 2)`.
- **Stars** only tell you whether that range crosses zero. Ignore them; read the range.
- "Not significant" ≠ "no effect." Distance: `r fmt(-0.08 - 2 * 0.06, 2)` to
  `r fmt(-0.08 + 2 * 0.06, 2)`. That's _uncertain_, not _zero_.
- Then go back to question 4: why is household income in this model?


# 7. How plausible was it before?

- Back to Week 1: **P( true | significant )** depends on the **base rate**.
- Warning signs:
  - The finding is surprising, cheap, and convenient.
  - Small sample, large effect.
  - Many outcomes measured, one reported.
  - Selected for publication or press (a collider). _(Week 4)_
- **Replications** and **pre-registration** raise the base rate.

::: {.fragment}
Extraordinary claims need more than one study.
:::


# Worked case: bicycles for girls {.center}

Muralidharan & Prakash (2017), _Cycling to School_, AEJ: Applied


## The claim

<!-- TODO: verify figures against the paper before presenting -->

- **Program**: Bihar gave girls entering secondary school money for a bicycle (2006–).
- **Finding**: girls' secondary enrollment rose substantially; the gender gap narrowed.
- Should Madhesh Province copy it?


## Seven questions, applied

| Question | Bihar bicycles |
|---|---|
| 1. Claim | <!-- TODO --> |
| 2. Measured | <!-- TODO --> |
| 3. Design | <!-- TODO: triple difference: girls vs. boys, Bihar vs. Jharkhand, before vs. after --> |
| 4. DAG / model | <!-- TODO --> |
| 5. Who | <!-- TODO --> |
| 6. Size / certainty | <!-- TODO --> |
| 7. Prior plausibility | <!-- TODO --> |
: {.nonincremental}

::: {.fragment}
Vote again: would you fund it in Madhesh? What would you need to know first?
:::


# Part 1: Summary

::: {.nonincremental}
1. **Claim**: descriptive, predictive or causal? Seeing or doing?
2. **Measure**: does the variable mean what the headline says?
3. **Design**: what decided who got X?
4. **DAG**: what did they control for, and should they have?
5. **Who**: whose data, and whose decision?
6. **Size & certainty**: in decision units, as a range.
7. **Prior**: how surprising, and how many tries?
:::


# Part 2: Asking your own question {.center}

Start from the decision, not the data


## Start with the decision

- What will you **do differently** depending on the answer?
- If the answer is high, you'll ___. If it's low, you'll ___.
- If both blanks say the same thing, **don't do the analysis.**

::: {.fragment}
"What do the data say about education?" is not a question. "Should we fund
transport for girls in grades 9–10 in this district?" is.
:::


## Building an estimand

::: {.callout-tip .nonincremental title="Estimand template"}
Among **[population]**, what is the **[summary]** of **[outcome]** after **[time]**,
if **[intervention]** compared to **[alternative]**?
:::

| Piece | Example |
|---|---|
| Population | Girls finishing grade 8 in rural Madhesh |
| Outcome | Enrolled in grade 9 |
| Intervention | Bicycle provided |
| Compared to what | No program (or: a cash transfer of the same value) |
| Time | One year later |
| Summary | Difference in share enrolled |
: {.nonincremental}

- The **comparison** is the part people forget. "Effect of X" always means "X
  compared to what?"

<!-- TODO: cite Lundberg, Johnson & Stewart (2021), "What Is Your Estimand?", ASR -->


## The target trial

**If you could run any experiment, what would it be?**

- Who would you randomize? To what? Measured when?
- This is the estimand made concrete.
- Then compare it with the data you **actually** have.
- Every gap between the two is a **limitation**, named before you see results.

| Ideal experiment | Data we have | Gap |
|---|---|---|
| <!-- TODO --> | <!-- TODO --> | <!-- TODO --> |
: {.nonincremental}

<!-- TODO: cite Hernán & Robins, target trial emulation -->


## From estimand to data

1. **Draw the DAG** for your estimand. _(Weeks 3–4)_
2. Which variables must you measure to close the back doors?
3. Can you measure them? If not, which **design** gets around them? _(Question 3)_
4. Who needs to be in the sample for the answer to apply to **your** decision?
   _(Question 5)_
5. How precise does the answer need to be to change the decision?


## Is it worth finding out?

- If every plausible answer leads to the same decision, the data has no value _to you_.
- If the decision is expensive and hard to reverse, precision is worth paying for.
- Sometimes the best decision is a **small pilot**, designed as an experiment.

<!-- TODO: simple "decision threshold" diagram: posterior vs. the cost-effectiveness line -->


# Part 2: Summary

- Start from the **decision**. If no answer would change it, stop.
- Write the **estimand**: population, outcome, intervention, comparison, time, summary.
- Imagine the **target trial**. Its gaps with your data are your limitations.
- The DAG says what to measure. The decision says how precise it needs to be.


# Part 3: Applied {.center}


## Round 1: Critique

<!-- TODO: 3-4 one-page study summaries, one per group -->

In groups, take one study summary and answer the seven questions.

- Which question was hardest to answer from what they reported?
- What's the **one thing** you'd need to know before acting?


## Round 2: Design

<!-- TODO: decision prompts, e.g. municipal budget decisions -->

Each group gets a decision. Write:

1. The decision and what each answer would change.
2. The estimand, using the template.
3. The target trial.
4. The data you could realistically get, and the gaps.


## Your project

- Rewrite your project's question using the **estimand template**.
- What is your **target trial**? How far is World Bank 2022 data from it?
- List your **scope limits**:
  - One year, cross-sectional
  - Country-level: what can you say about individuals?
  - Measurement: does each variable mean what its label says?
- Which of the seven questions would a skeptic ask you first?


# Homework

**A one-page critique memo** on a real report or finding from your own work.

- Organized by the seven questions.
- Ends with: **"What I'd need to see before acting on this."**

<!-- TODO: homework/week5.qmd with a worked example memo -->
