---
format:
  revealjs: default
---

```{r setup, include=FALSE}
library(tidyverse)
source("r/rlib.R")

knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 12,
  fig.height = 6.5,
  dpi = 150,
  dev.args = list(bg = "transparent")
)

BLUE <- "#1b91ff"
RED <- "#ff2c2d"
GRAY <- "#999999"
INK <- "#16314d"

theme_set(
  theme_minimal(base_size = 22) +
    theme(
      panel.grid.minor = element_blank(),
      plot.title = element_text(face = "bold", color = INK),
      legend.position = "top"
    )
)

fmt <- function(x, d = 1) formatC(x, format = "f", digits = d, big.mark = ",")
pct <- function(x) paste0(round(100 * x), "%")
```


# Data Lunch: Week 5 {.center}

Reading a finding, asking a question


# Should we act on this?

<!-- TODO: a real headline (ideally Nepal or South Asia) making a confident causal claim -->

> _"[Headline goes here]"_

- Hands up: would you fund a program based on this?
- Keep your answer. We'll vote again at the end of Part 1.


# Last four weeks → this week

- **Week 1**: P(A|B) ≠ P(B|A). Surprising findings are usually false.
- **Week 2**: Describe numbers by distributions. Coefficients have units.
- **Week 3**: Seeing ≠ doing. The DAG decides which number answers your question.
- **Week 4**: Forks, pipes, colliders. Holding the wrong thing fixed creates paradoxes.
- This week: put it together, from **both sides** of an analysis.


# Goals this week

- Part 1: Evaluating someone else's finding
  - Seven questions to ask of any claim
  - Worked case: bicycles for girls in Bihar
- Part 2: What's missing from the model?
  - Reading the DAG a study never drew
  - Missing common causes, mediators, and which way they bias the result
- Part 3: Asking your own question
  - From a decision to an estimand
  - The target trial: the experiment you wish you could run
- Part 4: Applied
  - Critique a study, design an estimand, revisit your project


# Part 1: Evaluating a finding {.center}

Seven questions


## Seven questions

::: {.nonincremental}
1. **What exactly is the claim?**
2. **What did they actually measure?**
3. **How was the data made?**
4. **What's in their DAG, and what's in their model?**
5. **Who is it about?**
6. **How big, and how sure?**
7. **How plausible was it before?**
:::

::: {.fragment}
Every one of these comes from a week you've already done, except one or two new pieces.
:::


# 1. What exactly is the claim?

::: {.columns}
::: {.column width="33%"}
::: {.callout-note .nonincremental title="Descriptive"}
_How much? How many?_

"38% of Kathmandu voters support the mayor."
:::
:::

::: {.column width="33%"}
::: {.callout-tip .nonincremental title="Predictive"}
_What will happen?_

"The mayor will probably win."
:::
:::

::: {.column width="33%"}
::: {.callout-important .nonincremental title="Causal"}
_What if we intervened?_

"Free school meals raise attendance."
:::
:::
:::

- Each kind needs different evidence. Causal claims need the most.
- Watch the verbs: _boosts, drives, leads to, protects against_ are causal claims, even
  when the study was only able to support a descriptive one.
- Exercise: rewrite the headline as an **estimand**. Is it seeing or doing? _(Week 3)_

<!-- TODO: 3-4 real headlines; students classify each and spot causal verbs on descriptive evidence -->


# 2. What did they actually measure?

- The **concept** in the headline vs. the **variable** in the data.
  - "Education" → years of schooling? enrollment? test scores?
  - "Women's college enrollment (% gross)": can exceed 100%. Why?
- **Proxies** are descendants: they carry the concept only partly. _(Week 4)_
- **Missing data**: who didn't answer, and why? Missingness is often selection on a collider.
- Measured once, or tracked over time? Self-reported, or observed?

<!-- TODO: example of a measure that drifts from its concept (e.g. "poverty" as income vs. consumption) -->


# 3. How was the data made?

::: {.columns}
::: {.column}
```{r obs-dag, fig.width = 7, fig.height = 4.5}
dag_plot(
  c("U -> X", "U -> Y", "X -> Y"),
  list(X = c(0, 0), U = c(1, 1.1), Y = c(2, 0)),
  labels = c(X = "Program", U = "Motivation, wealth, ...", Y = "Outcome"),
  above = "U",
  highlight = c("U -> X", "U -> Y"),
  exposure = "X",
  outcome = "Y",
  unobserved = "U"
)
```

**Observational**: people chose X. Everything that drove that choice is a possible
back door.
:::

::: {.column}
::: {.fragment}
```{r rct-dag, fig.width = 7, fig.height = 4.5}
dag_plot(
  c("R -> X", "U -> Y", "X -> Y"),
  list(X = c(0, 0), U = c(1, 1.1), Y = c(2, 0), R = c(-0.2, 1.1)),
  labels = c(R = "Coin flip", X = "Program", U = "Motivation, wealth, ...", Y = "Outcome"),
  above = c("U", "R"),
  exposure = "X",
  outcome = "Y",
  unobserved = "U"
)
```

**Randomized**: a coin flip chose X. Every arrow into X is **deleted**. This is
`do(X)` done physically.
:::
:::
:::


## Between the extremes

- **Randomized experiments**: strongest for "does it work _here, for these people_?"
- **Natural experiments**: something _like_ a coin flip (a lottery, a cutoff, a border,
  a policy rollout date).
- **Before / after, with a comparison group**: works if the groups would have moved
  together without the program.
- **Cross-sectional comparison**: one snapshot, many back doors. Everything in Week 4
  applies.

::: {.fragment}
Ask: **"What decided who got X?"** If the answer is "people chose," start drawing back
doors.
:::

<!-- TODO: regression to the mean slide here, or in Week 2: "we intervened on the worst
schools and they improved" (before/after with no comparison group) -->


# 4. What's in their DAG, and what's in their model?

Most reports never draw a DAG. Draw it for them.

::: {.nonincremental}
- Which **forks** did they control for? Which did they miss?
- Did they control for a **mediator**, hiding the effect? ("after adjusting for...")
- Did they select on a **collider**? ("among patients," "among graduates,"
  "among funded firms")
:::

::: {.fragment}
The phrase **"after controlling for..."** is the most important sentence in the
methods section. Check every variable in that list against your DAG. _(Week 4)_
:::

::: {.fragment}
This question gets **all of Part 2**.
:::


# 5. Who is it about?

- **Sample vs. population**: who was studied, and who are you making a decision for?
- **Would it work here?** Different place, time, prices, institutions, baseline rates.
- **Average vs. everyone**: an average effect can hide people it hurts.

::: {.fragment .callout-important .nonincremental title="Ecological fallacy"}
Countries with more X have more Y does **not** mean people with more X have more Y.

Your projects use country-level data. What can they say about individual women,
households, students?
:::

<!-- TODO: classic ecological example (e.g. Robinson 1950, literacy and immigration) -->


# 6. How big, and how sure?

::: {.columns}
::: {.column}
**How big**

- Translate into units that matter for the decision: rupees, students, lives per
  10,000.
- **Relative vs. absolute**: "doubles your risk" from 1 in 10,000 to 2 in 10,000 is
  one extra case per 10,000 people.
- Compare to the cost, and to the alternatives.
:::

::: {.column}
**How sure**

- Look at the **range** of plausible values, not just the middle.
- Does the decision change anywhere in that range?
- If the range runs from "large benefit" to "slight harm," you don't know yet. That is
  a finding too.
:::
:::


## Reading a table you're handed

Most reports you'll read aren't Bayesian. You can still read them.

<!-- TODO: replace with a real (or realistic) regression table from a report -->

| | Estimate | Std. error | |
|---|---:|---:|---|
| Program | 0.42 | 0.15 | \*\* |
| Household income (log) | 1.10 | 0.30 | \*\*\* |
| Distance to school (km) | −0.08 | 0.06 | |
: {.nonincremental}

- **Rough plausible range**: estimate ± 2 × std. error.
  - Program: `r fmt(0.42 - 2 * 0.15, 2)` to `r fmt(0.42 + 2 * 0.15, 2)`.
- **Stars** only tell you whether that range crosses zero. Ignore them; read the range.
- "Not significant" ≠ "no effect." Distance: `r fmt(-0.08 - 2 * 0.06, 2)` to
  `r fmt(-0.08 + 2 * 0.06, 2)`. That's _uncertain_, not _zero_.
- Then go back to question 4: why is household income in this model?


# 7. How plausible was it before?

- Back to Week 1: **P( true | significant )** depends on the **base rate**.
- Warning signs:
  - The finding is surprising, cheap, and convenient.
  - Small sample, large effect.
  - Many outcomes measured, one reported.
  - Selected for publication or press (a collider). _(Week 4)_
- **Replications** and **pre-registration** raise the base rate.

::: {.fragment}
Extraordinary claims need more than one study.
:::


# Worked case: bicycles for girls {.center}

Muralidharan & Prakash (2017), _Cycling to School_, AEJ: Applied


## The claim

<!-- TODO: verify figures against the paper before presenting -->

- **Program**: Bihar gave girls entering secondary school money for a bicycle (2006–).
- **Finding**: girls' secondary enrollment rose substantially; the gender gap narrowed.
- Should Madhesh Province copy it?


## Seven questions, applied

| Question | Bihar bicycles |
|---|---|
| 1. Claim | <!-- TODO --> |
| 2. Measured | <!-- TODO --> |
| 3. Design | <!-- TODO: triple difference: girls vs. boys, Bihar vs. Jharkhand, before vs. after --> |
| 4. DAG / model | <!-- TODO --> |
| 5. Who | <!-- TODO --> |
| 6. Size / certainty | <!-- TODO --> |
| 7. Prior plausibility | <!-- TODO --> |
: {.nonincremental}

::: {.fragment}
Vote again: would you fund it in Madhesh? What would you need to know first?
:::


# Part 1: Summary

::: {.nonincremental}
1. **Claim**: descriptive, predictive or causal? Seeing or doing?
2. **Measure**: does the variable mean what the headline says?
3. **Design**: what decided who got X?
4. **DAG**: what did they control for, and should they have?
5. **Who**: whose data, and whose decision?
6. **Size & certainty**: in decision units, as a range.
7. **Prior**: how surprising, and how many tries?
:::


# Part 2: What's missing from the model? {.center}

The critique you can make without any statistics


## The problem is upstream

::: {.nonincremental}
1. **Question**: what do we want to know?
2. **Causal model**: what causes what? (the DAG)
3. **Statistical model**: the regression, the priors, the code
4. **Estimate**: the numbers in the table
:::

- Methods sections and peer review spend almost all their time on steps 3 and 4.
- Step 2 is usually **never written down**. The list of controls stands in for it.
- A mistake in step 2 can't be fixed later. A bigger sample just gives you a more
  precise wrong answer.
- So you can challenge a finding without checking a single calculation: **what did
  their causal model leave out?**


## Their model implies a DAG

> _"Countries where more girls attend secondary school have fewer births per woman,
> after controlling for GDP per capita and contraceptive use."_

::: {.columns}
::: {.column}
```{r implied-dag, fig.width = 7, fig.height = 5}
dag_plot(
  c("G -> S", "G -> B", "S -> B", "C -> B"),
  list(S = c(0, 0), B = c(3, 0), G = c(1.5, 1.3), C = c(1.5, -1.3)),
  labels = c(
    S = "Schooling", B = "Births",
    G = "GDP per capita", C = "Contraceptive use"
  ),
  above = "G",
  left = "S",
  right = "B",
  exposure = "S",
  outcome = "B"
)
```
:::

::: {.column}
- Each control is a variable they thought mattered.
- Each **missing arrow** is a claim too: _"nothing else causes both schooling and
  births."_
- Leaving an arrow out is a stronger claim than drawing one, and it's never stated
  out loud.
:::
:::


## Four questions for every arrow

For the exposure X and the outcome Y, ask:

1. **Common causes**: what could cause *both* X and Y? Is it in the model? [→ an
   unclosed **fork**]{.blue}
2. **Mediators**: *how* would X affect Y? Is that channel in the control list? [→ a
   blocked **pipe**]{.blue}
3. **Common effects**: was the sample chosen by something X and Y both affect? [→ an
   opened **collider**]{.blue}
4. **Reverse arrows**: could Y cause X? [→ the arrow points the wrong way]{.blue}

::: {.fragment}
Then repeat for each control variable. Why is it in the model? Is it a fork, a pipe or
a collider?
:::


## The usual suspects

In development data, the same few common causes come up again and again.

| Suspect | Examples |
|---|---|
| **Wealth / SES** | income, assets, GDP. Nearly everything follows it. _(Week 4)_ |
| **Geography** | urban vs. rural, terrain, climate, distance to markets |
| **Institutions** | state capacity, corruption, colonial history, conflict |
| **Culture and norms** | religion, gender norms, caste, language |
| **Time** | everything has trended together since 1990 |
| **Selection** | who chose X? motivation, ability, need |
| **Data quality** | richer, better-run places measure and report more |
: {.nonincremental}

- Go down the list for every claim. Ask whether each one plausibly points into **both**
  X and Y.


## Missing common causes

::: {.columns}
::: {.column}
```{r missing-forks, fig.width = 7, fig.height = 5.5}
dag_plot(
  c("G -> S", "G -> B", "S -> B", "U -> S", "U -> B", "N -> S", "N -> B"),
  list(
    S = c(0, 0), B = c(4, 0), G = c(2, 1.3),
    U = c(0.4, -1.6), N = c(3.6, -1.6)
  ),
  labels = c(
    S = "Schooling", B = "Births",
    G = "GDP", U = "Urbanization", N = "Gender norms"
  ),
  above = "G",
  left = "S",
  right = "B",
  exposure = "S",
  outcome = "B",
  unobserved = c("U", "N"),
  highlight = c("U -> S", "U -> B", "N -> S", "N -> B")
)
```
:::

::: {.column}
- **Urbanization**: cities have more secondary schools *and* make children more
  expensive to raise.
- **Gender norms**: where girls are expected to marry early, they leave school *and*
  start having children sooner.
- Also: family-planning programs, child mortality, religion, conflict...
- None of these is GDP. Controlling for GDP leaves these back doors open.
:::
:::


## Which way would they push?

You can often tell the **direction** of the bias without any data.

| Arrow into X | Arrow into Y | The missing confounder adds... |
|:-:|:-:|---|
| \+ | \+ | a false **positive** association |
| − | − | a false **positive** association |
| \+ | − | a false **negative** association |
| − | \+ | a false **negative** association |
: {.nonincremental}

::: {.fragment}
- Urbanization: **+** schooling, **−** births → a false negative association.
- Gender norms (early marriage): **−** schooling, **+** births → also false negative.
- Both push the same way as the reported effect, so the real effect of schooling is
  probably **smaller** than reported. Probably not zero, but smaller.
:::


## Mediators: direct and indirect effects

::: {.columns}
::: {.column}
```{r mediators, fig.width = 7, fig.height = 5}
dag_plot(
  c("S -> M", "M -> B", "S -> C", "C -> B", "S -> B"),
  list(S = c(0, 0), B = c(3, 0), M = c(1.5, 1.3), C = c(1.5, -1.3)),
  labels = c(
    S = "Schooling", B = "Births",
    M = "Age at marriage", C = "Contraceptive use"
  ),
  above = "M",
  left = "S",
  right = "B",
  exposure = "S",
  outcome = "B",
  highlight = c("S -> M", "M -> B", "S -> C", "C -> B")
)
```
:::

::: {.column}
- Schooling can lower fertility **through** later marriage and more contraceptive use.
  These are **indirect effects**.
- Whatever is left over is the **direct effect**.
- **Total effect** = direct + all indirect paths.
- They controlled for contraceptive use, so they removed one of the channels. Their
  coefficient is _"the effect of schooling, not counting its effect through
  contraception."_
:::
:::


## Which effect answers the question?

- **Funding girls' schooling?** You want the **total** effect. Every channel counts,
  so don't control for mediators.
- **Understanding the mechanism?** Direct and indirect effects matter. *Does* schooling
  work through contraception? If so, maybe fund clinics instead.
- Papers often add controls "to be safe." If a control is a mediator, it quietly
  changes the question being answered.

::: {.fragment .callout-warning .nonincremental title="Controlling for a mediator can also open a collider"}
Anything that causes both contraceptive use and births (e.g. access to health services)
now makes a new back door: schooling → **contraception** ← health access → births.
Controlling for a mediator only gives the direct effect if nothing else causes both the
mediator and the outcome. _(Week 4)_
:::


## Writing the critique

A response you can give without re-running any analysis:

::: {.callout-tip .nonincremental title="Template"}
The authors' model assumes nothing causes both **[X]** and **[Y]** except **[their
controls]**. Plausible common causes it leaves out include **[U₁, U₂]**. These would
probably make the estimate **[too large / too small]**.

**[M]** is in the control list, but it's a channel through which X affects Y. So the
estimate is the effect of X *not through M*, which is not the same as **[the policy
question]**.
:::

- Name specific variables, and say **which way** each one would push.
- "There may be omitted variables" isn't a critique. It's true of every study.


## Your turn

For each claim: name two missing common causes, any mediators in the controls, and the
direction each would push.

1. Households with a toilet have healthier children, after controlling for income.
2. Countries with more women in parliament have lower maternal mortality, after
   controlling for GDP and health spending.
3. Districts with more microfinance lenders have more women in paid work, after
   controlling for population.
4. Households receiving remittances spend more on schooling, after controlling for
   household size and caste.


## Possible answers

1. **Forks**: mother's education, clean water source, urban location. Most of them
   raise both toilet ownership and child health, which makes toilets look **more**
   effective than they are.
2. **Mediator**: health spending. Women legislators may lower maternal mortality *by*
   raising it, so controlling for it hides part of the effect. **Forks**: democracy,
   gender norms, women's education.
3. **Forks**: roads and markets. **Selection**: if lenders target poorer districts, the
   effect looks **smaller**.
4. **Forks**: migrant households were already better off and better connected.
   **Mediator** to watch: if they had controlled for income, they'd block the main
   channel.

::: {.fragment}
None of these needed a single number. They needed a DAG.
:::


# Part 2: Summary

- Most findings never state their causal model. **Draw it for them** from the control
  list.
- For X → Y and each control, look for **common causes**, **mediators**, **common
  effects**, and **reverse arrows**.
- Go through **the usual suspects**: wealth, geography, institutions, norms, time,
  selection, data quality.
- **Sign the bias**: say which way each missing variable would push.
- Ask whether the question needs the **total** or the **direct** effect, and whether
  the controls match that.


# Part 3: Asking your own question {.center}

Start from the decision, not the data


## Start with the decision

- What will you **do differently** depending on the answer?
- If the answer is high, you'll ___. If it's low, you'll ___.
- If both blanks say the same thing, **don't do the analysis.**

::: {.fragment}
"What do the data say about education?" is not a question. "Should we fund
transport for girls in grades 9–10 in this district?" is.
:::


## Building an estimand

::: {.callout-tip .nonincremental title="Estimand template"}
Among **[population]**, what is the **[summary]** of **[outcome]** after **[time]**,
if **[intervention]** compared to **[alternative]**?
:::

| Piece | Example |
|---|---|
| Population | Girls finishing grade 8 in rural Madhesh |
| Outcome | Enrolled in grade 9 |
| Intervention | Bicycle provided |
| Compared to what | No program (or: a cash transfer of the same value) |
| Time | One year later |
| Summary | Difference in share enrolled |
: {.nonincremental}

- The **comparison** is the part people forget. "Effect of X" always means "X
  compared to what?"

<!-- TODO: cite Lundberg, Johnson & Stewart (2021), "What Is Your Estimand?", ASR -->


## The target trial

**If you could run any experiment, what would it be?**

- Who would you randomize? To what? Measured when?
- This is the estimand made concrete.
- Then compare it with the data you **actually** have.
- Every gap between the two is a **limitation**, named before you see results.

| Ideal experiment | Data we have | Gap |
|---|---|---|
| <!-- TODO --> | <!-- TODO --> | <!-- TODO --> |
: {.nonincremental}

<!-- TODO: cite Hernán & Robins, target trial emulation -->


## From estimand to data

1. **Draw the DAG** for your estimand. _(Weeks 3–4)_
2. Which variables must you measure to close the back doors?
3. Can you measure them? If not, which **design** gets around them? _(Question 3)_
4. Who needs to be in the sample for the answer to apply to **your** decision?
   _(Question 5)_
5. How precise does the answer need to be to change the decision?


## Is it worth finding out?

- If every plausible answer leads to the same decision, the data has no value _to you_.
- If the decision is expensive and hard to reverse, precision is worth paying for.
- Sometimes the best decision is a **small pilot**, designed as an experiment.

<!-- TODO: simple "decision threshold" diagram: posterior vs. the cost-effectiveness line -->


# Part 3: Summary

- Start from the **decision**. If no answer would change it, stop.
- Write the **estimand**: population, outcome, intervention, comparison, time, summary.
- Imagine the **target trial**. Its gaps with your data are your limitations.
- The DAG says what to measure. The decision says how precise it needs to be.


# Part 4: Applied {.center}


## Round 1: Critique

<!-- TODO: 3-4 one-page study summaries, one per group -->

In groups, take one study summary and answer the seven questions.

- Draw the DAG their control list implies. Add the **missing common causes** and
  **mediators**, and say which way each would push the estimate.

- Which question was hardest to answer from what they reported?
- What's the **one thing** you'd need to know before acting?


## Round 2: Design

<!-- TODO: decision prompts, e.g. municipal budget decisions -->

Each group gets a decision. Write:

1. The decision and what each answer would change.
2. The estimand, using the template.
3. The target trial.
4. The data you could realistically get, and the gaps.


## Your project

- Rewrite your project's question using the **estimand template**.
- Critique your own DAG as if a stranger wrote it: which **common causes** of your
  exposure and outcome aren't in it? Which way would each bias your estimate?
- Are any of your controls **mediators**? Do you want the total or the direct effect?
- What is your **target trial**? How far is World Bank 2022 data from it?
- List your **scope limits**:
  - One year, cross-sectional
  - Country-level: what can you say about individuals?
  - Measurement: does each variable mean what its label says?
- Which of the seven questions would a skeptic ask you first?


# Homework

**A one-page critique memo** on a real report or finding from your own work.

- Organized by the seven questions.
- Includes the DAG their model implies, the common causes and mediators it leaves out,
  and the direction of each bias.
- Ends with: **"What I'd need to see before acting on this."**

<!-- TODO: homework/week5.qmd with a worked example memo -->
