---
format: 
  revealjs: default
---

# Data Lunch: Week 3 {.center notes="notes/week1_notes.md#title"}

Models

::: {.footer}
The Open Institute For Social Science
:::

```{r setup, echo=FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
library(tidybayes)
library(bayesplot)
library(cmdstanr)
library(jsonlite)
library(rethinking)
options(mc.cores = parallel::detectCores())
```

# What have we done so far?
- We chose a parameter.
- We stated our *prior* belief.
- We made observations.
- We calculated the *likelihood* of the things we observed.
- We combined our prior and likelihood to produce an updated *posterior* belief.

# What is a model?

- Our **data** is the observation we made [(# of Surveys, # of Votes)]{.fragment .green}
- A **model** is a theoretical description of how this data was generated, combining:
  - Our *prior* belief about our parameter(s) [(Popularity)]{.fragment .blue}
  - A method of calculating our *likelihood* from data

# What is "fitting"?

- We "fit" our model to our data using Bayes' Theorem.
- This updates our prior -> posterior.
- We can use different techniques for this process:
  - Stones in a bag: we counted all possibilities
  - Election polling: we used "grid approximation" to estimate the parameter every 5%
  - Going forward: we'll use software (Markov Chain Montey Carlo, or MCMC)



# Distributions Demo {.no-title}

```{shinylive-r}
#| standalone: true
#| eval: false
{{< include shiny/distributions.R >}}
``` 


# Writing Priors
We need a way to *describe* our prior belief for all values of our parameter (Popularity)

::: {.columns}

::: {.column width="33%"}
::: {.fragment}

```{r prior1, echo=FALSE, warning=FALSE, message=FALSE}
x <- seq(0, 100, length.out = 1001)
y <- dunif(x, min = 0, max = 100)

plot(
  x,
  y,
  type = "l",
  lwd = 2,
  col = "red",
  xlim = c(0, 100),
  xlab = "Votes",
  ylab = "Density",
  main = "Popularity ~ Uniform(0,100)",
  cex.main = 3
)
```

:::

::: {.fragment}

```{r prior2, echo=FALSE, warning=FALSE, message=FALSE}
x <- seq(0, 100, length.out = 1001)
y <- dunif(x, min = 20, max = 80)

plot(
  x,
  y,
  type = "l",
  lwd = 2,
  col = "red",
  xlim = c(0, 100),
  xlab = "Votes",
  ylab = "Density",
  main = "Popularity ~ Uniform(20,80)",
  cex.main = 3
)
```

:::
:::

::: {.column width="33%"}
::: {.fragment}

```{r prior3, echo=FALSE, warning=FALSE, message=FALSE}
x <- seq(0, 100, length.out = 1001)
y <- dnorm(x, mean = 50, sd = 20)

plot(
  x,
  y,
  type = "l",
  lwd = 2,
  col = "red",
  xlim = c(0, 100),
  xlab = "Votes",
  ylab = "Density",
  main = "Popularity ~ Normal(50,20)",
  cex.main = 3
)
```

:::

::: {.fragment}

```{r prior4, echo=FALSE, warning=FALSE, message=FALSE}
x <- seq(0, 100, length.out = 1001)
y <- dnorm(x, mean = 40, sd = 5)

plot(
  x,
  y,
  type = "l",
  lwd = 2,
  col = "red",
  xlim = c(0, 100),
  xlab = "Votes",
  ylab = "Density",
  main = "Popularity ~ Normal(40,5)",
  cex.main = 3
)
```

:::
:::

::: {.column width="33%"}
::: {.fragment}

```{r prior5, echo=FALSE, warning=FALSE, message=FALSE}
x <- seq(0, 100, length.out = 1001)
y <- dnorm(x, mean = 60, sd = 15)
plot(
  x,
  y,
  type = "l",
  lwd = 2,
  col = "red",
  xlim = c(0, 100),
  xlab = "Votes",
  ylab = "Density",
  main = "Popularity ~ Normal(60,15)",
  cex.main = 3
)
```

:::

::: {.fragment}

```{r prior6, echo=FALSE, warning=FALSE, message=FALSE}
x <- seq(0, 100, length.out = 1001)
y <- dnorm(x, mean = 80, sd = 3)

plot(
  x,
  y,
  type = "l",
  lwd = 2,
  col = "red",
  xlim = c(0, 100),
  xlab = "Votes",
  ylab = "Density",
  main = "Popularity ~ Normal(80,3)",
  cex.main = 3
)
```

:::
:::

:::

# Writing likelihood functions

We need a way to describe how our data got made, linking our parameter (Popularity) to our observed data (Surveys and Votes)

$$
{\color{green}{Votes}} \sim \mathrm{Binomial}({\color{green}{Surveys}}, {\color{blue}{Popularity}})
$$

[Observed Data]{.green}

[Parameter(s)]{.blue}

# Writing complete models

$$
\begin{align}
\color{green}{Votes} &\sim \color{green}{\mathrm{Binomial}(Surveys, Popularity)} \\
\color{red}{Popularity} &\sim \color{red}{\mathrm{Normal}(60\%, 15\%)}
\end{align}
$$

[Likelihood function]{.green}

[Prior Belief]{.red}

<!-- # Two major types of models:

- Understanding the distribution of variables (parameter estimation)
  - Based on limited survey data, who do we think will win an election?
  - Based on social network analysis, how many Nepali citizens currently live abroad?

- Understanding relationships between variables (regression)
  - Does yoga lead to longer, healthier lives?
  - Does education make people happy?
  - What interventions can change national birth rates? -->


<!-- # Cause is super weird
- it's always a theoretical attribution, simplifying arbitrary complexity
- in the social science, we tend to think about cause as sites of intervention -->

# Understanding global birth rates

```{r wb_setup, echo=FALSE, warning=FALSE, message=FALSE}
filename <- "datasets/wb2022.csv"

# build a variable label vector for better labelling of df
label_vec <- read_csv(filename) |>
  distinct(`Series Code`, `Series Name`) |>
  filter(!is.na(`Series Code`), !is.na(`Series Name`)) |>
  group_by(`Series Code`) |> # in case of duplicates
  summarise(label = first(`Series Name`), .groups = "drop") |>
  with(setNames(label, `Series Code`))

# compose dataframe
df <- read_csv(filename) |>
  select(-c(`Series Name`)) |>
  slice(1:(n() - 5)) |>
  pivot_longer(
    cols = matches("^\\d{4} \\[YR\\d{4}\\]$"),
    names_to = "year",
    values_to = "value",
    names_transform = list(year = ~ str_extract(.x, "^\\d{4}"))
  ) |>
  pivot_wider(names_from = `Series Code`, values_from = value) |>
  mutate(across(
    any_of(names(label_vec)),
    ~ {
      attr(.x, "label") <- unname(label_vec[cur_column()])
      .x
    }
  )) |>
  left_join(
    read_csv("datasets/wb_country.csv"),
    join_by(`Country Code` == Code)
  )

# build variable summary showing only eligible variables
label_df <- df |>
  summarise(across(everything(), ~ sum(is.na(.x)))) |>
  pivot_longer(everything(), names_to = "column", values_to = "n_NA") |>
  # filter(n_NA < 50) |>
  rowwise() |>
  filter(
    !any(is.na(df[df$`Country Name` == "Nepal", .data$column, drop = TRUE]))
  ) |>
  ungroup() |>
  transmute(
    `Series Code` = column,
    label = unname(label_vec[column])
  ) |>
  filter(!is.na(label))
```

:::{.r-stack }

:::{.fragment width="1605"}
```{r wb_fertility1, echo=FALSE, warning=FALSE, message=FALSE, out.width="1605"}
ggplot(df, aes(x = SP.DYN.TFRT.IN)) +
  geom_histogram(bins = 11, colour = "darkgray") +
  scale_fill_discrete(
    labels = scales::label_wrap(20)
  ) +
  theme(legend.key.spacing.y = unit(0.2, 'cm'), legend.key = element_blank())
```



:::

:::{.fragment width="1605"}
```{r wb_fertility2, echo=FALSE, warning=FALSE, message=FALSE, out.width="1605"}
ggplot(df, aes(x = SP.DYN.TFRT.IN, fill = Region)) +
  geom_histogram(bins = 11, colour = "darkgray", position = "stack") +
  scale_fill_discrete(
    labels = scales::label_wrap(20)
  ) +
  theme(legend.key.spacing.y = unit(0.2, 'cm'), legend.key = element_blank())
```
:::


:::


# Birthrate ~ Women's College Enrollment

```{r wb_college, echo=FALSE, warning=FALSE, message=FALSE}
ggplot(df, aes(x = SE.TER.ENRR.FE, y = SP.DYN.TFRT.IN)) +
  geom_point(aes(color = Region)) +
  stat_smooth(method = "lm") +
  scale_color_discrete(
    labels = scales::label_wrap(20)
  ) +
  theme(legend.key.spacing.y = unit(0.2, 'cm'), legend.key = element_blank())
```


# Birthrate ~ Infant Mortality

:::{.r-stack}

:::{.fragment width="1605"}
```{r wb_infant1, echo=FALSE, warning=FALSE, message=FALSE, out.width="1605"}
ggplot(df, aes(x = SP.DYN.IMRT.IN, y = SP.DYN.TFRT.IN)) +
  geom_point(aes(color = Region)) +
  stat_smooth(method = "lm") +
  scale_color_discrete(
    labels = scales::label_wrap(20)
  ) +
  scale_x_continuous(limits = c(0, 100)) +
  theme(legend.key.spacing.y = unit(0.2, 'cm'), legend.key = element_blank())

```

:::

:::{.fragment width="100%"}

```{r wb_infant2, echo=FALSE, warning=FALSE, message=FALSE, out.width="1605"}
ggplot(df, aes(x = log(SP.DYN.IMRT.IN), y = SP.DYN.TFRT.IN)) +
  geom_point(aes(color = Region)) +
  stat_smooth(method = "lm") +
  scale_color_discrete(
    labels = scales::label_wrap(20)
  ) +
  # scale_x_continuous(limits=c(0,100)) +
  theme(legend.key.spacing.y = unit(0.2, 'cm'), legend.key = element_blank())

```

:::
:::

# Data: no meaning without a story

```{=html}
<iframe
  id="dag-frame-1"
  data-external="1"
  style="width:100%; aspect-ratio: 2 / 1; border:1px solid black;"
  src="https://dagitty.net/dags.html"
></iframe>
```


# Modeling individual weight

```{r weight, echo=FALSE, warning=FALSE, message=FALSE}
data(Howell1)

df_weight <- Howell1 |>
  filter(age >= 18) |>
  mutate(
    height_std = standardize(height),
    age_std = standardize(age),
    sex = ifelse((male == 0), "female", "male") |> as_factor()
  )

ggplot(df_weight, aes(x=weight)) +
  geom_histogram(bins=32, color="gray") + 
  geom_vline(
    xintercept = mean(df_weight$weight),
    color = "red",
    linetype = "dashed"
  )
```


# Vertical plot 
```{r weight_vert, echo=FALSE, warning=FALSE, message=FALSE}
ggplot(df_weight) +
  geom_jitter(
    data = df_weight,
    aes(x = 0, y = weight, color=sex),
    shape = "cross",
    color = "purple"
  ) +
  geom_hline(
    yintercept = mean(df_weight$weight),
    color = "red",
    linetype = "dashed"
  ) +
  xlim(-2, 2)
```

# Vertical plot (sex differentiated)

```{r weight_sex, echo=FALSE, warning=FALSE, message=FALSE}
ggplot(df_weight) +
  geom_jitter(
    data = df_weight,
    aes(x = ifelse(sex == "male", -1, 1), y = weight, color=sex),
    shape = "cross",
  ) +
  geom_hline(
    yintercept = mean((df_weight |> filter(sex == "male"))$weight),
    color = "red",
    linetype = "dashed"
  ) +
  geom_hline(
    yintercept = mean((df_weight |> filter(sex == "female"))$weight),
    color = "blue",
    linetype = "dashed"
  ) +
  xlim(-2, 2)
```

# Scatter plot (with height)
:::{.r-stack width="1605"}
:::{.fragment width="1605"}
```{r weight_height1, echo=FALSE, warning=FALSE, message=FALSE, out.width="1605"}
df_weight |>
  ggplot(aes(x = height, y = weight)) +
  geom_point(color = "purple", shape = "plus") +
  labs(x = "Height (cm)", y = "Weight (kg)")

```
:::
:::{.fragment width="1605"}
```{r weight_height2, echo=FALSE, warning=FALSE, message=FALSE, out.width="1605"}
df_weight |>
  ggplot(aes(x = height, y = weight)) +
  geom_point(color = "purple", shape = "plus") +
  geom_smooth(method="lm", se=FALSE) +
  labs(x = "Height (cm)", y = "Weight (kg)")

```
:::

:::{.fragment width="1605"}
```{r weight_height3, echo=FALSE, warning=FALSE, message=FALSE, out.width="1605"}
df_weight |>
  ggplot(aes(x = height / 2.56, y = weight)) +
  geom_point(color = "purple", shape = "plus") +
  geom_smooth(method="lm", se=FALSE) +
  labs(x = "Height (inches)", y = "Weight (kg)")

```
:::

:::{.fragment width="1605"}
```{r weight_height4, echo=FALSE, warning=FALSE, message=FALSE, out.width="1605"}
df_weight |>
  ggplot(aes(x = scale(height), y = weight)) +
  geom_point(color = "purple", shape = "plus") +
  geom_smooth(method="lm", se=FALSE) +
  labs(x = "Height (standardized)", y = "Weight (kg)")

```
:::

:::

# Simple stories

```{=html}
<iframe
  id="dag-frame-1"
  data-external="1"
  style="width:100%; aspect-ratio: 2 / 1; border:1px solid black;"
  src="https://dagitty.net/dags.html"
></iframe>
```


# Model definition (Weight)

$$
\begin{align}
weight_i &\sim \text{Normal}(\mu, \sigma) \\
\mu &\sim \text{Normal}(50, 10) \\
\sigma &\sim \text{Uniform}(0, 50)
\end{align}
$$

```{r model_weight, message=FALSE, results="hide"}
set.seed(101)
model <- ulam(
  alist(
    weight ~ normal(mu, sigma),
    mu ~ normal(50, 10),
    sigma ~ uniform(0, 50)
  ),
  data = df_weight
)
```
```{r precis_weight}
precis(model, prob = 0.95)
```


# Model (Weight ~ Sex)
$$
\begin{align}
\text{weight}_i &\sim \text{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha_{sex[i]} \\
\alpha_{sex} &\sim \text{Normal}(50, 10) \\
\sigma &\sim \text{Uniform}(0, 50)
\end{align}
$$

```{r model_weight2, message=FALSE, results="hide"}
set.seed(101)
model <- ulam(
  alist(
    weight ~ normal(mu, sigma),
    mu <- a[sex],
    a[sex] ~ normal(50, 10),
    sigma ~ uniform(0, 50)
  ),
  data = df_weight
)
```
```{r precis_weight2}
precis(model, prob = 0.95, depth=2)
```


# Model (Weight ~ Height)

$$
\begin{align}
\text{weight}_i &\sim \text{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta\times \text{height}_i \\
\alpha &\sim \text{Normal}(50, 10) \\
\beta &\sim \text{Normal}(0, 10) \\
\sigma &\sim \text{Uniform}(0, 50)
\end{align}
$$

```{r model_weight3, message=FALSE, results="hide"}
set.seed(101)
model <- ulam(
  alist(
    weight ~ normal(mu, sigma),
    mu <- a + b * height_std,
    a ~ normal(50, 10),
    b ~ normal(0, 10),
    sigma ~ uniform(0, 50)
  ),
  data = df_weight
)
```

```{r precis_weight3}
precis(model, prob = 0.95)
```

# Model (Weight ~ Sex + Height)

$$
\begin{align}
\text{weight}_i &\sim \text{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha_{sex[i]} + \beta\times \text{height}_i \\
\alpha_{sex[i]} &\sim \text{Normal}(50, 10) \\
\beta &\sim \text{Normal}(0, 10) \\
\sigma &\sim \text{Uniform}(0, 50)
\end{align}
$$

```{r model_weight4, message=FALSE, results="hide"}
set.seed(101)
model <- ulam(
  alist(
    weight ~ normal(mu, sigma),
    mu <- a[sex] + b * height_std,
    a[sex] ~ normal(50, 10),
    b ~ normal(0, 10),
    sigma ~ uniform(0, 50)
  ),
  data = df_weight
)
```

```{r precis_weight4}
precis(model, prob = 0.95, depth = 2)
```

# Model (Fertility ~ Women in College)
```{r model1a, message=FALSE, results="hide"}
model1a <- cmdstan_model("models/wb_model1a.stan", force_recompile = TRUE)
fit1a <- df |>
  transmute(
    fertility_rate = SP.DYN.TFRT.IN,
    women_in_college = scale(SE.TER.ENRR.FE) |> as.numeric()
  ) |>
  drop_na() |>
  compose_data() |>
  model1a$sample()
```

```{r model1a_output}
# fit1a |>
#   spread_draws(alpha, beta, sigma) |>
#   ggplot(aes(x = beta)) +
#   stat_halfeye()

# fit1a$summary()

fit1a$output()
```


# Model (Fertility ~ Infant Mortality)
```{r model1b, message=FALSE, results="hide"}
model1b <- cmdstan_model("models/wb_model1b.stan")
fit1b <- df |>
  transmute(
    fertility_rate = SP.DYN.TFRT.IN,
    infant_mortality = scale(SP.DYN.IMRT.IN) |> as.numeric()
  ) |>
  drop_na() |>
  compose_data() |>
  model1b$sample(chains=1)
```

```{r model1b_output}
# fit1b |>
#   spread_draws(alpha, beta, sigma) |>
#   ggplot(aes(x = beta)) +
#   stat_halfeye()

# fit1b$summary()
```


# Model (Fertility ~ Women in College + Infant Mortality)
```{r model2, message=FALSE, results="hide"}
model2 <- cmdstan_model("models/wb_model2.stan")
fit2 <- df |>
  transmute(
    fertility_rate = SP.DYN.TFRT.IN,
    women_in_college = scale(SE.TER.ENRR.FE) |> as.numeric(),
    infant_mortality = scale(SP.DYN.IMRT.IN) |> as.numeric()
  ) |>
  drop_na() |>
  compose_data() |>
  model2$sample(chains=1)

```

:::{.columns}
:::{.column}

```{r model2_output1, message=FALSE, results="hide"}
# fit2 |>
#   spread_draws(alpha, beta_college, beta_mortality, sigma) |>
#   ggplot() +
#   stat_halfeye(aes(x = beta_college))
```


:::
:::{.column}

```{r model2_output2, message=FALSE, results="hide"}
# fit2 |>
#   spread_draws(alpha, beta_college, beta_mortality, sigma) |>
#   ggplot() +
#   stat_halfeye(aes(x = beta_mortality))
```

:::
:::


```{r model2_output3}
# fit2$summary()
```