---
format: 
  revealjs: default
---

# Data Lunch: Week 4 {.center notes="notes/week1_notes.md#title"}

Confounding

::: {.footer}
The Open Institute For Social Science
:::

```{r setup, echo=FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
```


# Confounds

- Forks
- Pipes
- Colliders
- Descendant



# Fork

```{r fork_data, echo=FALSE, warning=FALSE, message=FALSE}
#| code-summary: "Show data generator code"
# A = age at marriage
# H = happiness
# G = geography
set.seed(116)
n <- 500
df <- tibble(
  G = rbinom(n, 1, 0.5),
  A = rnorm(n, 2 * G - 1),
  H = rnorm(n, 2 * G - 1)
)
```


## Happiness ~ Age at Marriage

::: {.r-stack}

:::{.fragment}
```{r}
#| fig-width: 14
#| fig-height: 8
#| out-width: 100%
plot(
  H ~ A,
  data = df,
  col = "#555",
  lwd = 2,
  xlab = "Age at Marriage",
  ylab = "Happiness"
)
```

:::

:::{.fragment}
```{r}
#| fig-width: 14
#| fig-height: 8
#| out-width: 100%
plot(
  H ~ A,
  data = df,
  col = "#555",
  lwd = 2,
  xlab = "Age at Marriage",
  ylab = "Happiness"
)
abline(lm(H ~ A, data = df), lwd = 3)
```

:::


:::{.fragment}
```{r}
#| fig-width: 14
#| fig-height: 8
#| out-width: 100%
plot(
  H ~ A,
  data = df,
  col = c("coral2", "cornflowerblue")[G + 1],
  lwd = 2,
  xlab = "Age at Marriage",
  ylab = "Happiness"
)
abline(lm(H ~ A, data = df), lwd = 3)
# abline(lm(H[G == 0] ~ A[G == 0], data = df), col = "coral2", lwd = 3)
# abline(lm(H[G == 1] ~ A[G == 1], data = df), col = "cornflowerblue", lwd = 3)
legend(
  "bottomright",
  legend = c("Rural", "Urban"),
  col = c("coral2", "cornflowerblue"),
  lwd = 2
)
```

:::

:::{.fragment}
```{r}
#| fig-width: 14
#| fig-height: 8
#| out-width: 100%
plot(
  H ~ A,
  data = df,
  col = c("coral2", "cornflowerblue")[G + 1],
  lwd = 2,
  xlab = "Age at Marriage",
  ylab = "Happiness"
)
abline(lm(H ~ A, data = df), lwd = 3)
abline(lm(H[G == 0] ~ A[G == 0], data = df), col = "coral2", lwd = 3)
abline(lm(H[G == 1] ~ A[G == 1], data = df), col = "cornflowerblue", lwd = 3)
legend(
  "bottomright",
  legend = c("Rural", "Urban"),
  col = c("coral2", "cornflowerblue"),
  lwd = 2
)
```

:::

:::


## Fork Graph

```{=html}
<iframe
  id="dag-frame-1"
  data-external="1"
  style="width:100%; aspect-ratio: 2 / 1; border:1px solid black;"
  src="https://dagitty.net/dags.html"
></iframe>
```






# Pipes

```{r pipe_data, echo=FALSE, warning=FALSE, message=FALSE}
#| code-summary: "Show data generator code"

# G = gender
# T = time spent studying
# R = exam results
set.seed(100)
n <- 500
df <- tibble(
  G = rbinom(n, 1, 0.5),
  T = rnorm(n, ifelse(G == 0, 2, -2), 2),
  R = rnorm(n, T, 3)
)
```

## Exam Results ~ Gender

::: {.r-stack}

:::{.fragment}
```{r}
#| fig-width: 14
#| fig-height: 8
#| out-width: 100%
plot(
  density(df$R),
  xlab = "Exam results",
  ylim = c(0, max(density(df$R)$y) * 1.5),
  main = "Exam Results Density Plot"
)
# lines(density(df$R[df$G == 0]), col = "blue")
# lines(density(df$R[df$G == 1]), col = "red")
legend(
  "topright",
  legend = c("Male", "Female"),
  col = c("blue", "red"),
  lwd = 2
)
```

:::

:::{.fragment}
```{r}
#| fig-width: 14
#| fig-height: 8
#| out-width: 100%
plot(
  density(df$R),
  xlab = "Exam results",
  ylim = c(0, max(density(df$R)$y) * 1.5),
  main = "Exam Results Density Plot"
)
lines(density(df$R[df$G == 0]), col = "blue")
lines(density(df$R[df$G == 1]), col = "red")
legend(
  "topright",
  legend = c("Male", "Female"),
  col = c("blue", "red"),
  lwd = 2
)
```

:::


:::

## Exam Results ~ Time Studying

::: {.r-stack}

:::{.fragment}
```{r}
#| fig-width: 14
#| fig-height: 8
#| out-width: 100%
plot(
  R ~ T,
  xlab = "Time Spent Studying",
  ylab = "Exam Results",
  data = df,
  # col = c(4, 2)[G + 1],
  lwd = 2
)
# legend(
#   "bottomright",
#   legend = c("Male", "Female"),
#   col = c("blue", "red"),
#   lwd = 2
# )
# abline(lm(R ~ T, data = df), lwd = 2)
# abline(lm(R[G == 0] ~ T[G == 0], data = df), col = 4, lwd = 2)
# abline(lm(R[G == 1] ~ T[G == 1], data = df), col = 2, lwd = 2)
```

:::

:::{.fragment}
```{r}
#| fig-width: 14
#| fig-height: 8
#| out-width: 100%
plot(
  R ~ T,
  xlab = "Time Spent Studying",
  ylab = "Exam Results",
  data = df,
  # col = c(4, 2)[G + 1],
  lwd = 2
)
# legend(
#   "bottomright",
#   legend = c("Male", "Female"),
#   col = c("blue", "red"),
#   lwd = 2
# )
abline(lm(R ~ T, data = df), lwd = 2)
# abline(lm(R[G == 0] ~ T[G == 0], data = df), col = 4, lwd = 2)
# abline(lm(R[G == 1] ~ T[G == 1], data = df), col = 2, lwd = 2)
```

:::

:::{.fragment}
```{r}
#| fig-width: 14
#| fig-height: 8
#| out-width: 100%
plot(
  R ~ T,
  xlab = "Time Spent Studying",
  ylab = "Exam Results",
  data = df,
  col = c(4, 2)[G + 1],
  lwd = 2
)
legend(
  "bottomright",
  legend = c("Male", "Female"),
  col = c("blue", "red"),
  lwd = 2
)
abline(lm(R ~ T, data = df), lwd = 2)
# abline(lm(R[G == 0] ~ T[G == 0], data = df), col = 4, lwd = 2)
# abline(lm(R[G == 1] ~ T[G == 1], data = df), col = 2, lwd = 2)
```

:::


:::{.fragment}
```{r}
#| fig-width: 14
#| fig-height: 8
#| out-width: 100%
plot(
  R ~ T,
  xlab = "Time Spent Studying",
  ylab = "Exam Results",
  data = df,
  col = c(4, 2)[G + 1],
  lwd = 2
)
legend(
  "bottomright",
  legend = c("Male", "Female"),
  col = c("blue", "red"),
  lwd = 2
)
abline(lm(R ~ T, data = df), lwd = 2)
abline(lm(R[G == 0] ~ T[G == 0], data = df), col = 4, lwd = 2)
abline(lm(R[G == 1] ~ T[G == 1], data = df), col = 2, lwd = 2)
```

:::




:::

## Pipe Graph

```{=html}
<iframe
  id="dag-frame-1"
  data-external="1"
  style="width:100%; aspect-ratio: 2 / 1; border:1px solid black;"
  src="https://dagitty.net/dags.html"
></iframe>
```



# Colliders

```{r collider_data, echo=FALSE, warning=FALSE, message=FALSE}
#| code-summary: "Show data generator code"
# F = family wealth
# H = hardwork
# J = prestigeous job

n <- 500
df <- tibble(
  F = rnorm(n),
  H = rnorm(n),
  R = rnorm(n, 0, 0.3),
  J = ifelse(F + H + R > 0.5, TRUE, FALSE)
)
```


## Hardwork ~ Family Wealth

:::{.r-stack}


:::{.fragment}
```{r}
#| fig-width: 14
#| fig-height: 8
#| out-width: 100%
plot(
  H ~ F,
  data = df,
  lwd = 2,
  col = ifelse(J, "royalblue", "transparent"),
  xlab = "Family Wealth",
  ylab = "Hardwork"
)
abline(lm(H[J == TRUE] ~ F[J == TRUE], data = df), lwd = 3, col = "royalblue")
```

:::


:::{.fragment}
```{r}
#| fig-width: 14
#| fig-height: 8
#| out-width: 100%
plot(
  H ~ F,
  data = df,
  lwd = 2,
  col = ifelse(J, "royalblue", "darkgray"),
  xlab = "Family Wealth",
  ylab = "Hardwork"
)
abline(lm(H[J] ~ F[J], data = df), lwd = 3, col = "royalblue")
abline(lm(H[!J] ~ F[!J], data = df), lwd = 3, col = "darkgray")
# abline(lm(H ~ F, data = df), lwd = 3, col = "black")
legend(
  "topright",
  legend = c("Job", "No Job"),
  col = c("blue", "darkgray"),
  lwd = 2
)
```

:::


:::{.fragment}
```{r}
#| fig-width: 14
#| fig-height: 8
#| out-width: 100%
plot(
  H ~ F,
  data = df,
  lwd = 2,
  col = ifelse(J, "royalblue", "darkgray"),
  xlab = "Family Wealth",
  ylab = "Hardwork"
)
abline(lm(H[J] ~ F[J], data = df), lwd = 3, col = "royalblue")
abline(lm(H[!J] ~ F[!J], data = df), lwd = 3, col = "darkgray")
abline(lm(H ~ F, data = df), lwd = 3, col = "black")
legend(
  "topright",
  legend = c("Job", "No Job"),
  col = c("blue", "darkgray"),
  lwd = 2
)
```

:::


:::

## Collider Graph

```{=html}
<iframe
  id="dag-frame-1"
  data-external="1"
  style="width:100%; aspect-ratio: 2 / 1; border:1px solid black;"
  src="https://dagitty.net/dags.html"
></iframe>
```


# Summary

| **Collider Type** | **Should you control?** |
|-------------------|------------------------:|
| Fork              |                Probably |
| Pipe              |              It depends |
| Collider          |            Probably Not |
| Descender         |         Depends on Type |


