---
format: 
  revealjs: default
---

# Data Lunch: Week 3 {.center notes="notes/week1_notes.md#title"}

Models

::: {.footer}
The Open Institute For Social Science
:::

# What have we done so far?
- We chose a parameter.
- We stated our *prior* belief.
- We made observations.
- We calculated the *likelihood* of the things we observed.
- We combined our prior and likelihood to produce an updated *posterior* belief.

# What is a model?

- A **model** is a theoretical description of how data is generated
- It is the combination of:
  - Our *priors*
  - A method of calculating our *likelihood* from data.

<!-- # The mechanics of "fitting"

- In our stones in a bag example, we tested literally every possible value of our parameter.
- In our voting estimation example, we used "grid approximation" to test our parameter on 5% increments.
- As our models become more complex, these approaches become impractical.
- Software to the rescue, but even the most complicated software is just looking for ways to count possibilities more efficiently. -->

# Distributions Demo {.no-title}

```{shinylive-r}
#| standalone: true
{{< include shiny/distributions.R >}}
``` 


# Writing Priors

::: {.columns}

::: {.column width="33%"}
::: {.fragment}

```{r echo=FALSE, warning=FALSE, message=FALSE}
x <- seq(0, 100, length.out = 1001)
y <- dunif(x, min = 0, max = 100)

plot(
  x,
  y,
  type = "l",
  lwd = 2,
  col = "red",
  xlim = c(0, 100),
  xlab = "Votes",
  ylab = "Density",
  main = "Popularity ~ Uniform(0,100)",
  cex.main = 3
)
```

:::

::: {.fragment}

```{r echo=FALSE, warning=FALSE, message=FALSE}
x <- seq(0, 100, length.out = 1001)
y <- dunif(x, min = 20, max = 80)

plot(
  x,
  y,
  type = "l",
  lwd = 2,
  col = "red",
  xlim = c(0, 100),
  xlab = "Votes",
  ylab = "Density",
  main = "Popularity ~ Uniform(20,80)",
  cex.main = 3
)
```

:::
:::

::: {.column width="33%"}
::: {.fragment}

```{r echo=FALSE, warning=FALSE, message=FALSE}
x <- seq(0, 100, length.out = 1001)
y <- dnorm(x, mean = 50, sd = 20)

plot(
  x,
  y,
  type = "l",
  lwd = 2,
  col = "red",
  xlim = c(0, 100),
  xlab = "Votes",
  ylab = "Density",
  main = "Popularity ~ Normal(50,20)",
  cex.main = 3
)
```

:::

::: {.fragment}

```{r echo=FALSE, warning=FALSE, message=FALSE}
x <- seq(0, 100, length.out = 1001)
y <- dnorm(x, mean = 40, sd = 5)

plot(
  x,
  y,
  type = "l",
  lwd = 2,
  col = "red",
  xlim = c(0, 100),
  xlab = "Votes",
  ylab = "Density",
  main = "Popularity ~ Normal(40,5)",
  cex.main = 3
)
```

:::
:::

::: {.column width="33%"}
::: {.fragment}

```{r echo=FALSE, warning=FALSE, message=FALSE}
x <- seq(0, 100, length.out = 1001)
y <- dnorm(x, mean = 60, sd = 15)
plot(
  x,
  y,
  type = "l",
  lwd = 2,
  col = "red",
  xlim = c(0, 100),
  xlab = "Votes",
  ylab = "Density",
  main = "Popularity ~ Normal(60,15)",
  cex.main = 3
)
```

:::

::: {.fragment}

```{r echo=FALSE, warning=FALSE, message=FALSE}
x <- seq(0, 100, length.out = 1001)
y <- dnorm(x, mean = 80, sd = 3)

plot(
  x,
  y,
  type = "l",
  lwd = 2,
  col = "red",
  xlim = c(0, 100),
  xlab = "Votes",
  ylab = "Density",
  main = "Popularity ~ Normal(80,3)",
  cex.main = 3
)
```

:::
:::

:::

# How to write likelihood functions

- How did our data get made?

$$
Votes \sim \mathrm{Binomial}(Surveys, Popularity)
$$



# Complete Model

$$
\begin{align}
Votes &\sim \mathrm{Binomial}(Surveys, Popularity) \\
Popularity &\sim \mathrm{Normal}(60\%, 15\%)
\end{align}
$$

<!-- # Two major types of models:

- Understanding the distribution of variables (parameter estimation)
  - Based on limited survey data, who do we think will win an election?
  - Based on social network analysis, how many Nepali citizens currently live abroad?

- Understanding relationships between variables (regression)
  - Does yoga lead to longer, healthier lives?
  - Does education make people happy?
  - What interventions can change national birth rates? -->


<!-- # Cause is super weird
- it's always a theoretical attribution, simplifying arbitrary complexity
- in the social science, we tend to think about cause as sites of intervention -->


# Modeling individual weight

```{r echo=FALSE, warning=FALSE, message=FALSE}
library(rethinking)
library(tidyverse)
data(Howell1)

df <- Howell1 |>
  filter(age >= 18) |>
  mutate(
    height_std = standardize(height),
    age_std = standardize(age),
    sex = ifelse((male == 0), "female", "male") |> as_factor()
  )

ggplot(df, aes(x=weight)) +
  geom_histogram(bins=32, color="gray") + 
  geom_vline(
    xintercept = mean(df$weight),
    color = "red",
    linetype = "dashed"
  )
```


# Vertical plot 
```{r echo=FALSE, warning=FALSE, message=FALSE}
ggplot(df) +
  geom_jitter(
    data = df,
    aes(x = 0, y = weight, color=sex),
    shape = "cross",
    color = "purple"
  ) +
  geom_hline(
    yintercept = mean(df$weight),
    color = "red",
    linetype = "dashed"
  ) +
  xlim(-2, 2)
```

# Vertical plot (sex differentiated)

```{r echo=FALSE, warning=FALSE, message=FALSE}
ggplot(df) +
  geom_jitter(
    data = df,
    aes(x = ifelse(sex == "male", -1, 1), y = weight, color=sex),
    shape = "cross",
  ) +
  geom_hline(
    yintercept = mean((df |> filter(sex == "male"))$weight),
    color = "red",
    linetype = "dashed"
  ) +
  geom_hline(
    yintercept = mean((df |> filter(sex == "female"))$weight),
    color = "blue",
    linetype = "dashed"
  ) +
  xlim(-2, 2)
```

# Scatter plot (with height)
:::{.r-stack}
:::{.fragment}
```{r echo=FALSE, warning=FALSE, message=FALSE}
df |>
  ggplot(aes(x = height, y = weight)) +
  geom_point(color = "purple", shape = "plus") +
  labs(x = "Height (cm)", y = "Weight (kg)")

```
:::
:::{.fragment}
```{r echo=FALSE, warning=FALSE, message=FALSE}
df |>
  ggplot(aes(x = height, y = weight)) +
  geom_point(color = "purple", shape = "plus") +
  geom_smooth(method="lm", se=FALSE) +
  labs(x = "Height (cm)", y = "Weight (kg)")

```
:::

:::{.fragment}
```{r echo=FALSE, warning=FALSE, message=FALSE}
df |>
  ggplot(aes(x = height / 2.56, y = weight)) +
  geom_point(color = "purple", shape = "plus") +
  geom_smooth(method="lm", se=FALSE) +
  labs(x = "Height (inches)", y = "Weight (kg)")

```
:::

:::{.fragment}
```{r echo=FALSE, warning=FALSE, message=FALSE}
df |>
  ggplot(aes(x = scale(height), y = weight)) +
  geom_point(color = "purple", shape = "plus") +
  geom_smooth(method="lm", se=FALSE) +
  labs(x = "Height (standardized)", y = "Weight (kg)")

```
:::

:::




# Model definition

$$
\begin{align}
weight_i &\sim \text{Normal}(\mu, \sigma) \\
\mu &\sim \text{Normal}(50, 10) \\
\sigma &\sim \text{Uniform}(0, 50)
\end{align}
$$

```{r, message=FALSE, results="hide"}
set.seed(101)
model <- ulam(
  alist(
    weight ~ normal(mu, sigma),
    mu ~ normal(50, 10),
    sigma ~ uniform(0, 50)
  ),
  data = df
)
```
```{r}
precis(model, prob = 0.95)
```


# Model (with sex)
$$
\begin{align}
\text{weight}_i &\sim \text{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha_{sex[i]} \\
\alpha_{sex[i]} &\sim \text{Normal}(50, 10) \\
\sigma &\sim \text{Uniform}(0, 50)
\end{align}
$$

```{r, message=FALSE, results="hide"}
set.seed(101)
model <- ulam(
  alist(
    weight ~ normal(mu, sigma),
    mu <- a[sex],
    a[sex] ~ normal(50, 10),
    sigma ~ uniform(0, 50)
  ),
  data = df
)
```
```{r}
precis(model, prob = 0.95, depth=2)
```


# Model (with height)

$$
\begin{align}
\text{weight}_i &\sim \text{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta\times \text{height}_i \\
\alpha &\sim \text{Normal}(50, 10) \\
\beta &\sim \text{Normal}(0, 10) \\
\sigma &\sim \text{Uniform}(0, 50)
\end{align}
$$

```{r, message=FALSE, results="hide"}
set.seed(101)
model <- ulam(
  alist(
    weight ~ normal(mu, sigma),
    mu <- a + b * height_std,
    a ~ normal(50, 10),
    b ~ normal(0, 10),
    sigma ~ uniform(0, 50)
  ),
  data = df
)
```

```{r}
precis(model, prob = 0.95)
```

# Model (Sex + Height)

$$
\begin{align}
\text{weight}_i &\sim \text{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha_{sex[i]} + \beta\times \text{height}_i \\
\alpha_{sex[i]} &\sim \text{Normal}(50, 10) \\
\beta &\sim \text{Normal}(0, 10) \\
\sigma &\sim \text{Uniform}(0, 50)
\end{align}
$$

```{r, message=FALSE, results="hide"}
set.seed(101)
model <- ulam(
  alist(
    weight ~ normal(mu, sigma),
    mu <- a[sex] + b * height_std,
    a[sex] ~ normal(50, 10),
    b ~ normal(0, 10),
    sigma ~ uniform(0, 50)
  ),
  data = df
)
```

```{r}
precis(model, prob = 0.95, depth = 2)
```