---
format:
  revealjs: default
---

```{r setup, include=FALSE}
library(tidyverse)
source("r/rlib.R")

knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 12,
  fig.height = 6.5,
  dpi = 150,
  dev.args = list(bg = "transparent")
)

BLUE <- "#1b91ff"
RED <- "#ff2c2d"
GRAY <- "#999999"
INK <- "#16314d"

theme_set(
  theme_minimal(base_size = 22) +
    theme(
      panel.grid.minor = element_blank(),
      plot.title = element_text(face = "bold", color = INK),
      legend.position = "top"
    )
)

fmt <- function(x, d = 1) formatC(x, format = "f", digits = d, big.mark = ",")
pct <- function(x) paste0(round(100 * x), "%")

# The three building blocks, drawn the same way everywhere
fork_dag <- function(x = "X", z = "Z", y = "Y", labels = NULL, direct = FALSE, ...) {
  dag_plot(
    c(paste(z, "->", x), paste(z, "->", y), if (direct) paste(x, "->", y)),
    setNames(list(c(0, 0), c(1, 1.1), c(2, 0)), c(x, z, y)),
    labels = labels, above = z, exposure = x, outcome = y, ...
  )
}
pipe_dag <- function(x = "X", z = "Z", y = "Y", labels = NULL, ...) {
  dag_plot(
    c(paste(x, "->", z), paste(z, "->", y)),
    setNames(list(c(0, 0), c(1, 1.1), c(2, 0)), c(x, z, y)),
    labels = labels, above = z, exposure = x, outcome = y, ...
  )
}
collider_dag <- function(x = "X", z = "Z", y = "Y", labels = NULL, ...) {
  dag_plot(
    c(paste(x, "->", z), paste(y, "->", z)),
    setNames(list(c(0, 1.1), c(1, 0), c(2, 1.1)), c(x, z, y)),
    labels = labels, above = c(x, y), exposure = x, outcome = y, ...
  )
}
```


# Data Lunch: Week 4 {.center}

Confounders and paradoxes


# Last week → this week

- The variables we put in a model can change, or even reverse, the answer.
- That's because effects **flow along paths** through the DAG.
- This week: the rules.
  - Which paths are **open**?
  - Which variables should we **hold fixed**, and which should we leave alone?


# Goals this week

- Part 1: Concepts
  - The three building blocks of every DAG: **fork**, **pipe**, **collider**
  - The paradoxes each one creates
  - "Should I control for this?"
- Part 2: Applied
  - Spot the confound
  - Check your own project's DAG


# Three building blocks

::: {.columns}
::: {.column width="33%"}
**Fork**: a common cause

```{r blocks-fork, fig.width = 5, fig.height = 3.6}
fork_dag()
```
:::

::: {.column width="33%"}
**Pipe**: a mediator

```{r blocks-pipe, fig.width = 5, fig.height = 3.6}
pipe_dag()
```
:::

::: {.column width="33%"}
**Collider**: a common effect

```{r blocks-collider, fig.width = 5, fig.height = 3.6}
collider_dag()
```
:::
:::

- Every path in every DAG is built from these three pieces.
- For each, we ask two questions:
  - Does association flow between X and Y through Z?
  - What happens when we **hold Z fixed**?


## What "holding Z fixed" means

People say it many ways. They all mean the same thing:

::: {.nonincremental}
- "controlling for Z"
- "adjusting for Z"
- "conditioning on Z"
- "stratifying by Z"
:::

In practice, any of these:

- **Adding Z to a regression** (as we did with height last week).
- **Splitting the data into groups** of Z and comparing within each group.
- **Selecting your sample** by Z: only hospital patients, only published studies, only
  people with jobs.

::: {.fragment}
That last one is easy to miss. Often the data was "controlled" before you ever saw it.
:::


# The fork: a common cause {.center}

X ← Z → Y


## Does marrying later make you happier?

```{r fork-data}
# Made-up world: urban people marry later and are happier,
# but age at marriage has no effect on happiness
set.seed(116)
n <- 500
marriage <- tibble(
  urban = rbinom(n, 1, 0.5),
  age = 25 + 3 * rnorm(n, 2 * urban - 1),
  happiness = 5 + 1.5 * rnorm(n, 2 * urban - 1),
  place = if_else(urban == 1, "Urban", "Rural")
)
```

::: {.columns}
::: {.column}
```{r fork-all, fig.width = 7, fig.height = 5.5}
ggplot(marriage, aes(age, happiness)) +
  geom_point(color = GRAY, alpha = 0.6, size = 2) +
  geom_smooth(method = "lm", se = FALSE, color = INK, linewidth = 1.5) +
  labs(x = "Age at marriage", y = "Happiness", title = "Everyone")
```
:::

::: {.column}
::: {.fragment}
```{r fork-split, fig.width = 7, fig.height = 5.5}
ggplot(marriage, aes(age, happiness, color = place)) +
  geom_point(alpha = 0.6, size = 2) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.5) +
  scale_color_manual(values = c(Rural = RED, Urban = BLUE), name = NULL) +
  labs(x = "Age at marriage", y = "Happiness", title = "Split by geography")
```
:::
:::
:::

- Overall: people who marry later are happier. Within rural *or* urban areas: no
  relationship at all.


## The fork: what happened?

::: {.columns}
::: {.column width="45%"}
```{r fork-dag, fig.width = 6, fig.height = 4.5}
fork_dag(
  "A", "G", "H",
  labels = c(A = "Age at marriage", G = "Geography", H = "Happiness"),
  highlight = c("G -> A", "G -> H")
)
```
:::

::: {.column width="55%"}
- Urban people marry later **and** (in this made-up world) are happier.
- Geography is a **confounder**: a common cause of X and Y.
- The path A ← G → H is **open**: it creates a correlation with no causal effect behind
  it.
- Holding geography fixed **closes** the path.
:::
:::

::: {.callout-tip .nonincremental title="Fork rule"}
Association flows through a fork. **Control for the common cause** to close the path.
:::


## Do mobile phones lower birth rates?

```{r phones-data}
wb_raw <- read_csv(
  "datasets/wb2022.csv",
  na = c("", ".."),
  col_types = cols(.default = "c"),
  locale = locale(encoding = "latin1")
) |>
  select(code = `Country Code`, series = `Series Code`, value = `2022 [YR2022]`) |>
  filter(series %in% c("IT.CEL.SETS.P2", "SP.DYN.TFRT.IN")) |>
  distinct(code, series, .keep_all = TRUE) |>
  mutate(value = as.numeric(value)) |>
  pivot_wider(names_from = series, values_from = value)

phones <- read_csv("datasets/wb_country.csv") |>
  select(code = Code, income = `Income group`) |>
  inner_join(wb_raw, by = "code") |>
  transmute(
    phones = IT.CEL.SETS.P2,
    fertility = SP.DYN.TFRT.IN,
    income = factor(
      income,
      levels = c("Low income", "Lower middle income", "Upper middle income", "High income")
    )
  ) |>
  drop_na()

r_phones <- cor(phones$phones, phones$fertility)
r_within <- phones |>
  group_by(income) |>
  summarise(r = cor(phones, fertility))
```

::: {.columns}
::: {.column}
```{r phones-all, fig.width = 7, fig.height = 5.5}
ggplot(phones, aes(phones, fertility)) +
  geom_point(color = GRAY, alpha = 0.7, size = 2.5) +
  geom_smooth(method = "lm", se = FALSE, color = INK, linewidth = 1.5) +
  labs(
    x = "Mobile subscriptions per 100 people",
    y = "Births per woman",
    title = paste0("All countries: r = ", fmt(r_phones, 2))
  )
```
:::

::: {.column}
::: {.fragment}
```{r phones-split, fig.width = 7, fig.height = 5.5}
income_colors <- c("#ff2c2d", "#e69f00", "#009e73", "#1b91ff")

ggplot(phones, aes(phones, fertility, color = income)) +
  geom_point(alpha = 0.7, size = 2.5) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.5) +
  scale_color_manual(values = income_colors, name = NULL) +
  guides(color = guide_legend(nrow = 2)) +
  labs(
    x = "Mobile subscriptions per 100 people",
    y = "Births per woman",
    title = "Within income groups"
  ) +
  theme(legend.text = element_text(size = 14))
```
:::
:::
:::

- Real World Bank data, 2022. Within income groups the correlation mostly fades
  (*r* between `r fmt(min(r_within$r), 2)` and `r fmt(max(r_within$r), 2)`).
  **Wealth** drives both phones and smaller families.


## In development data, wealth is the fork behind everything

```{r fork-wealth, fig.width = 12, fig.height = 4.2}
dag_plot(
  c("W -> P", "W -> S", "W -> B", "W -> H", "W -> C", "S -> B"),
  list(
    W = c(2.6, 1.4), P = c(0, 0), S = c(1.3, 0), B = c(2.6, 0), H = c(3.9, 0), C = c(5.2, 0)
  ),
  labels = c(
    W = "Wealth / SES", P = "Phones", S = "Schooling", B = "Births",
    H = "Health", C = "Clean water"
  ),
  above = "W",
  exposure = "S",
  outcome = "B",
  highlight = c("W -> S", "W -> B")
)
```

- Your claim is **A → B**. Fine. But nearly always, **money → A** and **money → B** as well.
- Schooling, health, sanitation, phones, democracy, fertility, life expectancy: richer
  countries and households have more of the good ones and less of the bad ones.
- So any two development indicators will be correlated, whether or not one causes the
  other. **Assume wealth is a confounder until you've argued it isn't.**
- It's hard to control for completely. GDP and income group are rough
  [proxies](#descendants) for wealth, so some of the fork usually stays open.


## You've already met forks

::: {.columns}
::: {.column}
```{r fork-houses, fig.width = 6, fig.height = 4.2}
fork_dag(
  "B", "A", "P",
  labels = c(B = "Bedrooms", A = "Living area", P = "Price"),
  direct = TRUE,
  highlight = c("A -> B", "A -> P")
)
```

**Week 2:** more bedrooms looked valuable, because bigger houses have more of them.
:::

::: {.column}
```{r fork-surgeon, fig.width = 6, fig.height = 4.2}
fork_dag(
  "S", "C", "O",
  labels = c(S = "Surgeon", C = "Case difficulty", O = "Outcome"),
  direct = TRUE,
  highlight = c("C -> S", "C -> O")
)
```

**Week 3:** the famous surgeon looked worse, because he got the hardest cases.
:::
:::

- The classic **Simpson's paradox**, where a trend reverses when you split the data,
  is usually a fork.


## More forks

What's the hidden common cause?

- Ice cream sales ↔ drownings [→ hot weather]{.fragment .blue}
- Children's shoe size ↔ reading ability [→ age]{.fragment .blue}
- Firefighters at a fire ↔ damage done [→ size of the fire]{.fragment .blue}
- Carrying a lighter ↔ lung cancer [→ smoking]{.fragment .blue}
- Hospital visits ↔ dying [→ being ill]{.fragment .blue}


# The pipe: a mediator {.center}

X → Z → Y


## Is there a gender gap in exam results?

```{r pipe-data}
# Made-up world: one group studies more, and studying improves results.
# Gender has no effect on results except through study time.
set.seed(100)
n <- 500
exams <- tibble(
  male = rbinom(n, 1, 0.5),
  study = pmax(0, rnorm(n, if_else(male == 0, 13, 7), 3)),
  result = pmin(100, rnorm(n, 45 + 2 * study, 8)),
  gender = if_else(male == 1, "Male", "Female")
)
```

::: {.columns}
::: {.column}
```{r pipe-density, fig.width = 7, fig.height = 5.5}
ggplot(exams, aes(result, fill = gender)) +
  geom_density(alpha = 0.6, color = NA) +
  scale_fill_manual(values = c(Female = RED, Male = BLUE), name = NULL) +
  labs(x = "Exam result", y = NULL, title = "Results by gender") +
  theme(axis.text.y = element_blank())
```
:::

::: {.column}
::: {.fragment}
```{r pipe-split, fig.width = 7, fig.height = 5.5}
ggplot(exams, aes(study, result, color = gender)) +
  geom_point(alpha = 0.6, size = 2) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.5) +
  scale_color_manual(values = c(Female = RED, Male = BLUE), name = NULL) +
  labs(x = "Hours studied per week", y = "Exam result", title = "Holding study time fixed")
```
:::
:::
:::

- A clear gap between groups. But at the same study time, the two lines sit on top of
  each other: the gap disappears.


## The pipe: what happened?

::: {.columns}
::: {.column width="45%"}
```{r pipe-dag, fig.width = 6, fig.height = 4.5}
pipe_dag(
  "G", "T", "R",
  labels = c(G = "Gender", T = "Study time", R = "Exam result"),
  highlight = c("G -> T", "T -> R")
)
```
:::

::: {.column width="55%"}
- In this made-up world, gender affects results **only through** study time.
- The path G → T → R is **open**, and it is a real causal effect.
- Holding study time fixed **closes** the path, so the effect vanishes.
- Does gender matter? Yes, **through** study time.
:::
:::

::: {.callout-tip .nonincremental title="Pipe rule"}
Association flows through a pipe. Holding the mediator fixed **blocks the effect**. Only
do it if you want the *direct* effect.
:::


## Does the anti-fungal treatment work?

```{r fungus-data}
# After McElreath, Statistical Rethinking ch. 6: treatment prevents fungus,
# fungus slows growth. The treatment works, entirely through the fungus.
set.seed(71)
n <- 400
plants <- tibble(
  treated = rep(c("Untreated", "Treated"), each = n / 2),
  fungus = rbinom(n, 1, if_else(treated == "Treated", 0.1, 0.5)),
  growth = rnorm(n, 5 - 3 * fungus, 1)
)

growth_all <- plants |>
  group_by(treated) |>
  summarise(growth = mean(growth)) |>
  mutate(group = "All plants")

growth_split <- plants |>
  mutate(group = if_else(fungus == 1, "With fungus", "No fungus")) |>
  group_by(group, treated) |>
  summarise(growth = mean(growth), .groups = "drop")
```

::: {.columns}
::: {.column width="55%"}
```{r fungus-plot, fig.width = 7.5, fig.height = 5.5}
bind_rows(growth_all, growth_split) |>
  mutate(
    group = factor(group, levels = c("All plants", "No fungus", "With fungus")),
    treated = factor(treated, levels = c("Untreated", "Treated"))
  ) |>
  ggplot(aes(treated, growth, fill = treated)) +
  geom_col(width = 0.7) +
  facet_wrap(~group) +
  scale_fill_manual(values = c(Untreated = GRAY, Treated = BLUE), guide = "none") +
  labs(x = NULL, y = "Average growth (cm)") +
  theme(strip.text = element_text(size = 16, face = "bold"), axis.text.x = element_text(size = 14))
```
:::

::: {.column width="45%"}
```{r fungus-dag, fig.width = 6, fig.height = 3.8}
pipe_dag(
  "T", "F", "G",
  labels = c(T = "Treatment", F = "Fungus", G = "Growth"),
  highlight = c("T -> F", "F -> G")
)
```

- Overall, treated plants grow more.
- Compare plants with the **same fungus status** and the treatment looks useless.
:::
:::

- **Post-treatment bias**: never hold fixed something the treatment itself changes, unless
  that's the question.


## A real paradox: Berkeley admissions, 1973

```{r berkeley}
berkeley <- read_csv("datasets/berkeley.csv")

berk_overall <- berkeley |>
  group_by(sex) |>
  summarise(rate = mean(admitted), applicants = n())

berk_dept <- berkeley |>
  group_by(dept, sex) |>
  summarise(rate = mean(admitted), applicants = n(), .groups = "drop")

women_higher <- berk_dept |>
  select(dept, sex, rate) |>
  pivot_wider(names_from = sex, values_from = rate) |>
  summarise(n = sum(female > male)) |>
  pull(n)
```

::: {.columns}
::: {.column width="55%"}
```{r berkeley-plot, fig.width = 7.5, fig.height = 6}
bind_rows(
  berk_overall |> mutate(dept = "All"),
  berk_dept
) |>
  mutate(dept = fct_relevel(factor(dept), "All")) |>
  ggplot(aes(rate, fct_rev(dept), color = sex)) +
  geom_line(aes(group = dept), color = GRAY, linewidth = 1) +
  geom_point(aes(size = applicants), alpha = 0.9) +
  scale_color_manual(values = c(female = RED, male = BLUE), name = NULL) +
  scale_size_area(max_size = 10, guide = "none") +
  scale_x_continuous(labels = scales::label_percent()) +
  labs(x = "Admission rate", y = "Department")
```
:::

::: {.column width="45%"}
- Overall: **`r pct(berk_overall$rate[berk_overall$sex == "male"])`** of men admitted,
  **`r pct(berk_overall$rate[berk_overall$sex == "female"])`** of women.
- But in `r women_higher` of 6 departments, women were admitted at a **higher** rate.
- Women mostly applied to the departments that were **hardest to get into**.
- Dot size: number of applicants.
:::
:::


## Berkeley: which answer is right?

::: {.columns}
::: {.column width="45%"}
```{r berkeley-dag, fig.width = 6, fig.height = 4.5}
dag_plot(
  c("S -> D", "D -> A", "S -> A"),
  list(S = c(0, 0), D = c(1, 1.1), A = c(2, 0)),
  labels = c(S = "Sex", D = "Department", A = "Admitted"),
  above = "D",
  exposure = "S",
  outcome = "A",
  highlight = c("S -> D", "D -> A")
)
```
:::

::: {.column width="55%"}
- Department is a **pipe**: sex affects which department you apply to, and departments
  differ in how hard they are to get into.
- "Do admissions committees treat women differently?" → hold department **fixed**
  (the direct path S → A).
- "Are women less likely to get in overall?" → **don't** (the total effect). Then ask
  why women ended up applying to the most competitive departments.
:::
:::


# The collider: a common effect {.center}

X → Z ← Y


## Are rich kids lazy?

```{r collider-data}
# Made-up world: family wealth and hard work are unrelated,
# and either can get you a prestigious job
set.seed(7)
n <- 500
jobs <- tibble(
  wealth = rnorm(n),
  work = rnorm(n),
  job = wealth + work + rnorm(n, 0, 0.3) > 0.5,
  status = if_else(job, "Prestigious job", "No prestigious job")
)
```

::: {.columns}
::: {.column}
```{r collider-selected, fig.width = 7, fig.height = 5.5}
jobs |>
  filter(job) |>
  ggplot(aes(wealth, work)) +
  geom_point(color = BLUE, alpha = 0.7, size = 2) +
  geom_smooth(method = "lm", se = FALSE, color = BLUE, linewidth = 1.5) +
  coord_cartesian(xlim = range(jobs$wealth), ylim = range(jobs$work)) +
  labs(x = "Family wealth", y = "Hard work", title = "People with prestigious jobs")
```
:::

::: {.column}
::: {.fragment}
```{r collider-all, fig.width = 7, fig.height = 5.5}
ggplot(jobs, aes(wealth, work)) +
  geom_point(aes(color = status), alpha = 0.6, size = 2) +
  geom_smooth(method = "lm", se = FALSE, color = INK, linewidth = 1.5) +
  scale_color_manual(
    values = c("Prestigious job" = BLUE, "No prestigious job" = GRAY),
    name = NULL
  ) +
  labs(x = "Family wealth", y = "Hard work", title = "Everyone")
```
:::
:::
:::

- Among people with prestigious jobs, the richer ones seem to work less.
- But in the whole population (right), wealth and hard work are **unrelated**.


## The collider: what happened?

::: {.columns}
::: {.column width="45%"}
```{r collider-dag, fig.width = 6, fig.height = 4.5}
collider_dag(
  "W", "J", "H",
  labels = c(W = "Family wealth", J = "Prestigious job", H = "Hard work"),
  highlight = c("W -> J", "H -> J")
)
```
:::

::: {.column width="55%"}
- Wealth **or** hard work can get you the job.
- If someone has the job but isn't rich, they probably worked hard. If they didn't work
  hard, they're probably rich.
- The path W → J ← H is **closed** by default: no association.
- Holding J fixed, including by **only looking at people with jobs**, **opens** it.
:::
:::

::: {.callout-important .nonincremental title="Collider rule"}
No association flows through a collider, until you hold it fixed. **Don't control for
a common effect**, and beware of samples selected on one.
:::


## Why is published science so often wrong?

```{r grants}
# After McElreath, Statistical Rethinking ch. 6
set.seed(1914)
n <- 200
grants <- tibble(
  newsworthy = rnorm(n),
  trustworthy = rnorm(n),
  score = newsworthy + trustworthy,
  funded = score >= quantile(score, 0.9)
)
r_funded <- with(filter(grants, funded), cor(newsworthy, trustworthy))
```

::: {.columns}
::: {.column width="55%"}
```{r grants-plot, fig.width = 7, fig.height = 6}
ggplot(grants, aes(newsworthy, trustworthy, color = funded)) +
  geom_point(size = 3, alpha = 0.8) +
  geom_smooth(
    data = filter(grants, funded),
    method = "lm", se = FALSE, color = BLUE, linewidth = 1.5
  ) +
  scale_color_manual(values = c(`TRUE` = BLUE, `FALSE` = GRAY), guide = "none") +
  labs(x = "Newsworthiness", y = "Trustworthiness")
```
:::

::: {.column width="45%"}
- 200 grant proposals. Newsworthiness and trustworthiness are **unrelated**.
- Reviewers fund the top 10% on *both combined*.
- Among [funded]{.blue} studies: *r* = **`r fmt(r_funded, 2)`**.
- The most exciting published findings are the least trustworthy, with no bad intent
  from anyone.
- Remember Week 1: *most published research findings are false*.
:::
:::


## More colliders

What has been selected on?

- Restaurants in bad locations serve better food [→ only the good ones survive]{.fragment .blue}
- Among professional basketball players, height barely predicts success [→ short players had to be exceptional to make it]{.fragment .blue}
- Among hospital patients, two unrelated diseases look negatively related [→ either one can put you in hospital (Berkson's paradox)]{.fragment .blue}
- Among low-birth-weight babies, those of smoking mothers die *less* often [→ birth weight is caused by smoking *and* by other, more dangerous problems]{.fragment .blue}
- Early COVID studies found fewer smokers among hospitalized patients [→ hospital testing and admission were the collider (Griffith et al., 2020)]{.fragment .blue}


# The descendant {.center}

A variable caused by Z acts like a weaker copy of Z.


## Descendants

::: {.columns}
::: {.column}
```{r desc-fork, fig.width = 7, fig.height = 4.5}
dag_plot(
  c("Z -> X", "Z -> Y", "Z -> D"),
  list(X = c(0, 0), Z = c(1, 1.1), Y = c(2, 0), D = c(1, -0.4)),
  labels = c(X = "Phones", Z = "Wealth", Y = "Births", D = "Income group"),
  above = "Z",
  exposure = "X",
  outcome = "Y",
  unobserved = "Z",
  highlight = "Z -> D"
)
```

**A proxy for a confounder.** We can't measure wealth exactly, so we used income group.
It closes the fork *partly*, which is why some correlation was left over.
:::

::: {.column}
```{r desc-collider, fig.width = 7, fig.height = 4.5}
dag_plot(
  c("X -> Z", "Y -> Z", "Z -> D"),
  list(X = c(0, 1.1), Z = c(1, 0.3), Y = c(2, 1.1), D = c(2, -0.6)),
  labels = c(X = "Wealth", Z = "Job", Y = "Hard work", D = "Salary"),
  above = c("X", "Y"),
  exposure = "X",
  outcome = "Y",
  highlight = "Z -> D"
)
```

**A consequence of a collider.** Studying only high earners partly selects on having
the job, so it partly opens the collider.
:::
:::


# Should I control for it?

| Structure | Through Z, by default | If you hold Z fixed | Advice |
|-----------|-----------------------|---------------------|--------|
| **Fork** X ← Z → Y | open | closed | Control for it |
| **Pipe** X → Z → Y | open | closed | Only for the *direct* effect |
| **Collider** X → Z ← Y | closed | **open** | Don't |
| **Descendant** of Z | like Z, weaker | like Z, weaker | Same as Z |
: {.nonincremental}


## The back-door rule

To estimate the effect of X on Y:

1. List every path between X and Y.
2. **Close** every path that starts with an arrow *into* X (a "back door").
3. **Don't** hold fixed anything that X causes (if you want the total effect).
4. **Don't** hold fixed a collider, unless you also close the path it opens.

::: {.fragment}
Last week's examples follow this rule. Height was caused by sex, so we left it out
(rule 3). Living area was a back door into bedrooms, so we held it fixed (rule 2).
:::


## A paradox inside a paradox

::: {.columns}
::: {.column width="45%"}
```{r berkeley-u, fig.width = 6, fig.height = 4.8}
dag_plot(
  c("S -> D", "D -> A", "S -> A", "U -> D", "U -> A"),
  list(S = c(0, 0), D = c(1, 1.1), A = c(2, 0), U = c(2, 1.1)),
  labels = c(S = "Sex", D = "Department", A = "Admitted", U = "Ability"),
  above = c("D", "U"),
  exposure = "S",
  outcome = "A",
  unobserved = "U",
  highlight = c("S -> D", "U -> D", "U -> A")
)
```
:::

::: {.column width="55%"}
- Back to Berkeley. What if applicants' **ability** affects both which department they
  apply to and whether they get in?
- Now department is a pipe **and** a collider (S → D ← U).
- Holding department fixed opens S → D ← U → A, so even the "direct effect" is biased.
- No statistical trick fixes this. We need **better data** (a measure of ability) or
  **more science**.
:::
:::


# Part 1: Summary

- **Fork** (common cause): creates false associations. Control for it.
- **Pipe** (mediator): carries real effects. Controlling for it hides them.
- **Collider** (common effect): closed until you control for it, or select on it.
- **Descendants** act like weaker copies of their parents.
- Every paradox this week came from holding the wrong thing fixed, or failing to hold
  the right thing fixed.


# Part 2: Spot the confound {.center}


## Your turn

For each claim: draw the DAG, name the structure, and decide what to control for.

1. Students who attend tuition classes score *lower* on the SEE.
2. After adjusting for blood pressure, a new drug doesn't reduce heart attacks.
3. Among startups that got funding, those with weaker founders had better ideas.
4. Countries with more doctors per person have more cancer.
5. People who own more books live longer.
6. Among people admitted to hospital with COVID, smokers seemed to do better.


## Possible answers

1. **Fork**: students who are already struggling are the ones sent to tuition.
   Control for earlier results.
2. **Pipe**: the drug works *by* lowering blood pressure. Don't adjust if you want the
   total effect.
3. **Collider**: funding needs a strong founder *or* a strong idea. Don't trust patterns
   within a selected sample.
4. **Fork** (and a pipe): wealth → more doctors, and wealth → longer lives → more
   cancer. More doctors also *detect* more cancer.
5. **Fork**: education and income cause both.
6. **Collider**: hospitalization. Smoking and severe COVID can each put you there.

::: {.fragment}
There is often more than one defensible DAG. Arguing about which one is right is the
science.
:::


## Your project

- Open your project's `dag` in [dagitty.net](https://dagitty.net/dags.html) (next slide).
- Mark your **exposure** and **outcome**.
- List every path between them. Label each fork, pipe and collider.
- DAGitty's **"Causal effect identification"** panel lists the *adjustment sets*: the
  variables that close every back door without opening any collider.
- Do your model's variables match? If not, which DAG are you really assuming?


## DAGitty {.no-title}

```{=html}
<iframe
  data-external="1"
  style="width:100%; aspect-ratio: 2 / 1; border:1px solid black;"
  src="https://dagitty.net/dags.html"
></iframe>
```
