---
title: A Bayesian Workflow
format:
  html:
    df-print: paged
    include-in-header:
      - text: |
          <script src="js/dagitty.js"></script>
---

This document walks through a complete Bayesian workflow, from the first question to a
result we can defend.

Imagine you're in charge of admissions at a large university. You want to know if your
admissions process is fair. You've got data from thousands of individual applicants. What now?

The workflow has eight steps:

1. **Define the estimand.** State exactly what you want to know before you touch the data.
2. **Draw the causal model (DAG).** Write down what you assume causes what, based on what you know about the domain.
3. **Simulate the population.** Write a simulation that could produce data like yours.
4. **Write the formal model.** Derive the statistical model from the DAG, in math notation.
5. **Write the Stan code.** Translate the formal model into a program.
6. **Test on simulated data.** Check that the priors imply plausible data and that the model recovers parameters we already know.
7. **Fit and validate the model.** Run the sampler, check its diagnostics, and run posterior predictive checks. Can we trust the model?
8. **Compute the estimand and test its robustness.** What is the result, and how much does it depend on our assumptions?

The steps are numbered, but the workflow is a loop. If the model can't recover known
parameters in step 6, or can't reproduce the data in step 7, go back and fix the model,
the simulation, or the DAG. Each of those failures teaches us something.

```{r setup, message=FALSE, results="hide"}
# libraries
library(tidyverse)
library(cmdstanr)
library(tidybayes)
library(posterior)
library(bayesplot)

{{< include r/rlib.R >}}

# print formatting
options(scipen = 999)
set.seed(1)

# data import
df <- read.csv("datasets/berkeley.csv") |>
  mutate(
    sex = factor(sex, levels = c("male", "female")),
    dept = factor(dept)
  )
```

# Step 1: Define the estimand

The estimand is the precise quantity we want to learn. "Is admissions fair?" is too vague
to estimate. Before we write any code, we need to turn it into a number.

Suppose we find that women are admitted at a lower rate overall. That gap could come from
two places:

- A **direct** effect: departments treat women differently from men who apply to the same department.
- An **indirect** effect: women apply to different departments, and those departments admit fewer people.

Both are worth knowing about, but they call for different fixes. As the admissions
office, we control the decisions departments make, so we'll focus on the direct effect:

> **Estimand:** the average difference in admission probability between women and men
> who apply to the *same* department, averaged across departments in proportion to how many
> people apply to each.

We'll measure it as a difference in admission rates. For example, a result of −0.05
would mean women are admitted 5 percentage points less often than men who apply to the
same department.
Settling this first tells us what the model has to include and what we'll compute from it
at the end.

# Step 2: Draw the causal model (DAG)

A causal model contains no numbers. It's a picture of the variables and of which ones
cause which. You'll hear it called a DAG (Directed Acyclic Graph), which is a complicated
name for a simple idea.

<div class="dagitty" style="width: 100%; padding: 20px 0; aspect-ratio: 2/1;">
dag {
Admitted [outcome,pos=\"-0.619,0.077\"]
Department [exposure,pos=\"-0.858,0.192\"]
Sex [exposure,pos=\"-1.059,-0.272\"]
Department -> Admitted
Sex -> Admitted
Sex -> Department
}

</div>

<script>
DAGitty.setup()
</script>

How do you pick these variables? You use what you know about the domain. We include sex
because sex-based discrimination is a major social issue. We include department because
some departments are much harder to get into than others.

Each arrow is an assumption. Each missing arrow is an assumption too, and usually a
stronger one. The arrow $\text{Sex} \to \text{Admitted}$ is the direct effect. The path
$\text{Sex} \to \text{Department} \to \text{Admitted}$ is the indirect effect. The DAG
tells us that to estimate the direct effect we need to **condition on department**, so
department has to go into the model.

The DAG also leaves things out. Many other factors affect admission, like an applicant's
qualifications. Leaving them off the graph is fine as long as they aren't related to sex.
In other words, we're assuming that the men and women who apply are equally qualified on
average. That's a strong assumption, and we'll come back to it in Step 8.

# Step 3: Simulate the population

Before writing a statistical model, we write a *generative* one: a simulation that follows
the DAG and produces fake applicants. This makes us state precisely how we think the data
come about. Later, it gives us data where we know the right answer.

```{r simulation}
depts <- levels(df$dept)

# how selective each department is (baseline admission rates)
dept_rates <- c(A = 0.65, B = 0.60, C = 0.35, D = 0.33, E = 0.25, F = 0.07)

# where men and women tend to apply (Sex -> Department)
male_pref <- c(0.30, 0.20, 0.12, 0.15, 0.07, 0.14)
female_pref <- c(0.06, 0.02, 0.32, 0.20, 0.21, 0.19)

simulate_admissions <- function(
  n = 4500,
  discrimination = 0, # direct effect of being female, in log-odds (Sex -> Admitted)
  u_on_admit = 0, # how much qualifications matter for admission (not in the DAG)
  female_selection = 0 # how much more qualified female applicants are, in SDs (for Step 8)
) {
  tibble(
    sex = sample(c("male", "female"), n, replace = TRUE, prob = c(0.6, 0.4))
  ) |>
    mutate(
      u = rnorm(n, mean = female_selection * (sex == "female")),
      dept = if_else(
        sex == "female",
        sample(depts, n, replace = TRUE, prob = female_pref),
        sample(depts, n, replace = TRUE, prob = male_pref)
      ),
      p = plogis(
        qlogis(dept_rates[dept]) +
          discrimination * (sex == "female") +
          u_on_admit * u
      ),
      admitted = rbinom(n, 1, p),
      sex = factor(sex, levels = c("male", "female")),
      dept = factor(dept, levels = depts)
    ) |>
    select(dept, sex, u, admitted)
}

# the true value of the estimand in a simulated population: switch each applicant's sex,
# holding their department and qualifications fixed
true_direct_effect <- function(sim, discrimination, u_on_admit = 0) {
  sim |>
    mutate(
      base = qlogis(dept_rates[as.character(dept)]) + u_on_admit * u,
      diff = plogis(base + discrimination) - plogis(base)
    ) |>
    group_by(dept) |>
    summarise(n = n(), diff = mean(diff)) |>
    summarise(sum(n / sum(n) * diff)) |>
    pull()
}
```

Let's simulate a world with **no discrimination at all**: within every department, men
and women are admitted at the same rate. Women just tend to apply to more selective
departments.

```{r simulation_demo}
sim_fair <- simulate_admissions(discrimination = 0)

# overall admission rates by sex
sim_fair |>
  group_by(sex) |>
  summarise(applicants = n(), admit_rate = mean(admitted))
```

Women are admitted much less often overall, even though the simulation contains no
discrimination. The whole gap comes from the indirect path through department. This
is why the estimand has to be defined within departments, and why department belongs in
the model. If we compared overall rates, we'd find a gap that isn't there.

# Step 4: Write the formal model

Next, we turn the DAG into a statistical model:

$$
\begin{align}
\text{admitted}_i &\sim \mathrm{Bernoulli}(p_i) \\
\mathrm{logit}(p_i)
&= \alpha
+ \beta_{\mathrm{sex}[i]}
+ \gamma_{\mathrm{dept}[i]} \\
\alpha &\sim \mathrm{Normal}(0, 1.5) \\
\beta_{\mathrm{sex}} &\sim \mathrm{Normal}(0, 1) \\
\gamma_{\mathrm{dept}} &\sim \mathrm{Normal}(0, 1) \\
0 &= \textstyle\sum_{\mathrm{sex}} \beta = \sum_{\mathrm{dept}} \gamma
\end{align}
$$

It looks complicated, but it has four parts:

1. A **likelihood** (line 1): how randomness enters. Each applicant is admitted or not.
2. A **linear predictor** (line 2): how the arrows in the DAG combine. There's a baseline
   $\alpha$, a sex effect $\beta$, and a department effect $\gamma$.
3. **Priors** (lines 3–5): what we believe before we see the data.
4. **Constraints** (line 6): each set of effects adds up to zero. Without this, the
   model couldn't tell its parts apart. Adding 1 to $\alpha$ and subtracting 1 from every
   $\gamma$ would give exactly the same predictions. With the constraint, $\alpha$ is the
   overall average and every other term is a deviation from it.

Each term in the model matches a piece of the DAG. Sex and department both point into
admission, so both appear in the linear predictor. The estimand from Step 1 comes from
$\beta$, the term that lets women's and men's chances differ within a department.

The model also makes a simplifying assumption: the sex effect $\beta$ is the same in
every department. That keeps the model small. We'll find out in Step 7 whether the data
go along with it.

# Step 5: Write the Stan code

Now we translate the formal model into a program the computer can run:

```{.stan include="models/berkeley.stan" filename="models/berkeley.stan" code-line-numbers="true"}
```

A few details:

- The priors are written directly in the `model` block, exactly as in the formal model.
- Stan enforces the constraints with `sum_to_zero_vector`. Forcing
  effects to sum to zero shrinks their spread, so the code widens each prior's scale to
  compensate. That way, each effect's prior matches the formal model.
- The `prior_only` switch skips the likelihood. With it on, the model samples from the
  priors alone, which is what we need for prior predictive checks.
- `generated quantities` reports the admission probability for every sex × department
  cell. We'll compute the estimand from these.

We compile the model and write a small helper so every fit in this document uses the same
settings:

```{r compile, message=FALSE, results="hide"}
berkeley <- cmdstan_model("models/berkeley.stan", dir = ".stancache")

fit_model <- function(data, prior_only = 0, model = berkeley) {
  data |>
    select(dept, sex, admitted) |>
    compose_data() |>
    c(list(prior_only = prior_only)) |>
    model$sample(seed = 2026, refresh = 0, show_messages = FALSE)
}

# the estimand from Step 1, computed for every posterior draw
direct_effect <- function(fit, data = df) {
  weights <- data |>
    count(dept) |>
    mutate(w = n / sum(n)) |>
    pull(w)

  fit |>
    spread_draws(p[sex, dept]) |>
    mutate(sex = levels(data$sex)[sex]) |>
    pivot_wider(names_from = sex, values_from = p) |>
    mutate(w = weights[dept]) |>
    group_by(.draw) |>
    summarise(direct_effect = sum(w * (female - male)))
}
```

# Step 6: Test on simulated data

We test the model on data where we already know the answer before we trust it on data
where we don't. There are two checks.

### Prior predictive check

First, do the priors imply plausible data? We run the model with `prior_only = 1` and
look at the admission rates it expects before seeing any data.

```{r prior_predictive, message=FALSE, results="hide"}
fit_prior <- fit_model(df, prior_only = 1)
```

```{r prior_predictive_plot}
fit_prior |>
  spread_draws(p[sex, dept]) |>
  ggplot(aes(x = p)) +
  geom_histogram(bins = 40, fill = "#1b91ff", color = "white") +
  labs(
    x = "Admission probability for a sex × department cell",
    y = "Prior draws",
    title = "What the priors imply about admission rates"
  )

direct_effect(fit_prior) |>
  ggplot(aes(x = direct_effect)) +
  stat_halfeye(fill = "#1b91ff") +
  labs(
    x = "Direct effect (female − male)",
    y = NULL,
    title = "Prior on the estimand"
  )
```

The prior on the direct effect is centered on zero, so it doesn't assume discrimination
in either direction. It's also wide: gaps of 50 percentage points or more either way are allowed.

The admission rates show a problem. The priors pile up near 0% and 100%. That
happens because three Normal priors add up on the log-odds scale, and a log-odds of ±3 is
already 5% or 95%. These priors are looser than our actual knowledge: we don't believe
departments admit nearly everyone or nearly no one. This is a judgment call. We could
tighten $\alpha$ and $\gamma$ now. With thousands of applicants, we expect the data to
outweigh the priors, so we'll keep them and check that expectation directly in Step 8.

### Parameter recovery

Second, can the model find an effect we put there ourselves? We simulate three
populations: one with no discrimination, one where women are penalized by 0.5 on the
log-odds scale, and one where they're penalized by 1. We fit the model to each and check
whether the posterior covers the true value of the estimand.

```{r recovery, message=FALSE, results="hide"}
recovery <- tibble(discrimination = c(0, -0.5, -1)) |>
  mutate(
    sim = map(discrimination, \(b) simulate_admissions(discrimination = b)),
    truth = map2_dbl(sim, discrimination, true_direct_effect),
    estimate = map(sim, \(s) direct_effect(fit_model(s), data = s))
  )
```

```{r recovery_plot}
recovery |>
  select(discrimination, truth, estimate) |>
  unnest(estimate) |>
  ggplot(aes(x = direct_effect, y = factor(discrimination))) +
  stat_halfeye(fill = "#1b91ff") +
  geom_point(
    data = distinct(recovery, discrimination, truth),
    aes(x = truth),
    color = "red",
    shape = 4,
    size = 4,
    stroke = 1.5
  ) +
  labs(
    x = "Direct effect (female − male)",
    y = "Discrimination built into the simulation (log-odds)",
    title = "Can the model recover a known effect?",
    subtitle = "Red ×: the true value of the estimand"
  )
```

In each case, the true value falls inside the posterior. The model and the estimand code
work, at least in a world that matches our assumptions. If they didn't, we'd have a bug
or an identification problem, and we'd fix it here, where we know the right answer. Any
single simulated dataset is noisy, so a more thorough version of this check repeats it
many times with random true values. That's called *simulation-based calibration*.

# Step 7: Fit and validate the model

Now we fit the model to the real data.

```{r fitting, message=FALSE, results="hide"}
fit <- fit_model(df)
```

### Sampler diagnostics

Before we look at any results, we check whether the sampler worked. Divergent transitions
mean the sampler hit geometry it couldn't explore. $\hat{R}$ should be very close to 1,
which means the chains agree with each other. Effective sample size (ESS) should be in
the hundreds or more.

```{r diagnostics}
fit$diagnostic_summary()

fit$draws() |>
  subset_draws(variable = c("alpha", "beta", "gamma")) |>
  summarise_draws() |>
  replace_indexes(df, beta[sex], gamma[dept]) |>
  print(n = Inf, width = Inf)

mcmc_trace(fit$draws(c("alpha", "beta", "gamma")))
```

The trace plots should look like "fuzzy caterpillars": chains that overlap and wander
around the same region with no trends.

### Posterior predictive check

Can the fitted model reproduce the data it was fit to? For each posterior draw, we
simulate a new set of admissions for every sex × department cell and compare the result
to what we actually observed.

```{r ppc}
cells <- df |>
  group_by(sex, dept) |>
  summarise(applicants = n(), observed = mean(admitted), .groups = "drop")

fit |>
  spread_draws(p[sex, dept]) |>
  mutate(
    sex = factor(levels(df$sex)[sex], levels = levels(df$sex)),
    dept = factor(levels(df$dept)[dept], levels = levels(df$dept))
  ) |>
  left_join(cells, by = c("sex", "dept")) |>
  mutate(predicted = rbinom(n(), applicants, p) / applicants) |>
  ggplot(aes(x = dept, y = predicted, color = sex)) +
  stat_pointinterval(
    position = position_dodge(width = 0.5),
    .width = c(0.5, 0.9)
  ) +
  geom_point(
    data = cells,
    aes(y = observed),
    shape = 4,
    size = 3,
    stroke = 1.2,
    position = position_dodge(width = 0.5)
  ) +
  labs(
    x = "Department",
    y = "Admission rate",
    title = "Posterior predictive check",
    subtitle = "Intervals: model predictions. ×: observed rates."
  )
```

For most departments, the observed rates (×) fall inside the predicted intervals.
Department A is the exception. Women there are admitted 82% of the time, well above
anything the model predicts. The model can't capture this because in Step 4 we assumed the
sex effect is the same in every department, and department A breaks that pattern.

This is the posterior predictive check doing its job. It tells us exactly where the model
is too simple. In a full analysis, we'd go back to Step 4 and let the sex effect vary by
department (an *interaction*). For this demonstration we'll keep the simpler model, but
we'll keep the misfit in mind: our estimate is an average over departments that don't all
behave alike.

# Step 8: Compute the estimand and test its robustness

Now we can answer the question from Step 1. We compute the direct effect for every
posterior draw:

```{r estimand}
de <- direct_effect(fit)

de |>
  ggplot(aes(x = direct_effect)) +
  stat_halfeye(fill = "#1b91ff") +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(
    x = "Direct effect (female − male)",
    y = NULL,
    title = "Posterior for the estimand"
  )

de |>
  summarise(
    median = median(direct_effect),
    lower_90 = quantile(direct_effect, 0.05),
    upper_90 = quantile(direct_effect, 0.95),
    prob_women_disadvantaged = mean(direct_effect < 0)
  )
```

Among applicants to the same department, there's no sign that women are admitted less
often than men. The best estimate is that they're admitted about 2 percentage points
*more* often. The 90% interval runs from about −1 to +4 points, so the data are
consistent with a small advantage for women or no difference at all. The large overall
gap comes from *where* women
apply, not from how departments treat them. That's the pattern our Step 3 simulation
produced in a world with no discrimination.

A result is only as good as the assumptions behind it, so we test how much it depends on
them.

### Sensitivity to priors

We refit with a much tighter and a much looser prior on the sex effect. Each variant is
the same model with one line changed: the scale in the prior on `beta`.

```{r sensitivity_priors, message=FALSE, results="hide"}
# a copy of the model with the prior on beta set to Normal(0, beta_sd)
with_beta_prior <- function(beta_sd) {
  code <- readLines("models/berkeley.stan")
  line <- grep("beta ~ normal(0, 1 *", code, fixed = TRUE)
  stopifnot(length(line) == 1)
  code[line] <- sub(
    "normal(0, 1 *",
    paste0("normal(0, ", beta_sd, " *"),
    code[line],
    fixed = TRUE
  )
  cmdstan_model(write_stan_file(code, dir = ".stancache"))
}

prior_scenarios <- list(
  "Tight (beta ~ Normal(0, 0.1))" = with_beta_prior(0.1),
  "Default (beta ~ Normal(0, 1))" = berkeley,
  "Loose (beta ~ Normal(0, 2.5))" = with_beta_prior(2.5)
)

prior_sensitivity <- prior_scenarios |>
  imap(\(model, name) {
    direct_effect(fit_model(df, model = model)) |> mutate(scenario = name)
  }) |>
  list_rbind()
```

```{r sensitivity_priors_plot}
prior_sensitivity |>
  ggplot(aes(x = direct_effect, y = scenario)) +
  stat_halfeye(fill = "#1b91ff") +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(
    x = "Direct effect (female − male)",
    y = NULL,
    title = "Sensitivity to priors"
  )
```

The three posteriors are almost identical. Even a prior that strongly insists sex
doesn't matter barely moves the estimate. This confirms what we expected in Step 6: with
thousands of applicants, the data outweigh the overly wide priors. With a small sample,
this check would matter much more.

### Sensitivity to the DAG: an unobserved confounder

The strongest assumption we made is that the factors left out of the DAG aren't related
to sex: men and women who apply are equally qualified. Suppose that's wrong. Suppose women face barriers
*before* they apply, so only the most qualified women apply at all. Then, within any
department, the women are more qualified than the men. Comparing within departments would
make women look favored even if departments discriminate against them.

We can see how much this matters by going back to the Step 3 simulation. We build in real
discrimination (−0.5 log-odds), make qualifications matter for admission, and vary how
much more qualified the female applicants are:

```{r sensitivity_confounding, message=FALSE, results="hide"}
confounding <- tibble(female_selection = c(0, 0.5, 1)) |>
  mutate(
    sim = map(female_selection, \(k) {
      simulate_admissions(
        discrimination = -0.5,
        u_on_admit = 1,
        female_selection = k
      )
    }),
    truth = map_dbl(sim, \(s) {
      true_direct_effect(s, discrimination = -0.5, u_on_admit = 1)
    }),
    estimate = map(sim, \(s) direct_effect(fit_model(s), data = s))
  )
```

```{r sensitivity_confounding_plot}
confounding |>
  select(female_selection, truth, estimate) |>
  unnest(estimate) |>
  ggplot(aes(x = direct_effect, y = factor(female_selection))) +
  stat_halfeye(fill = "#1b91ff") +
  geom_point(
    data = distinct(confounding, female_selection, truth),
    aes(x = truth),
    color = "red",
    shape = 4,
    size = 4,
    stroke = 1.5
  ) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  geom_vline(
    xintercept = median(de$direct_effect),
    color = "#1b91ff",
    linetype = "dotted"
  ) +
  labs(
    x = "Estimated direct effect (female − male)",
    y = "How much more qualified female applicants are (SD)",
    title = "Real discrimination, hidden by an unobserved confounder",
    subtitle = "Red ×: true effect. Dotted line: our estimate from the real data."
  )
```

In every one of these simulations, departments discriminate against women (red ×). When
applicants are equally qualified, the model finds the discrimination. If women applicants
are half a standard deviation more qualified, the estimate shrinks to less than half the
true effect. At a full standard deviation, women look clearly *favored*, by even more than
our real-data estimate (dotted line).

So our real-data result, "no direct disadvantage for women", is credible only to the
extent that we believe male and female applicants are similarly qualified. The model
can't settle that question. It's an argument about the world, and settling it would take
more data, such as measures of applicant qualifications.

### The result

Within departments, we find no evidence that women are disadvantaged. If anything, they're
admitted about 2 percentage points more often. The conclusion holds up across a wide range
of priors. It comes with two caveats. First, the model averages over departments, and
department A doesn't fit that average well. Second, it rests on an assumption we can't test with this data: that male and
female applicants are similarly qualified. The
overall gap is real. It runs through department choice, so the place to look is why
women apply to more selective departments.
